---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Travis-CI Build Status](https://travis-ci.org/samdchamberlain/salinity_study.svg?branch=master)](https://travis-ci.org/samdchamberlain/salinity_study)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Salinity study

This package was used the generate the manuscript "Effect of drought-induced salinization on wetland methane emissions, gross
  ecosystem photosynthesis, and their interactions." by Samuel D. Chamberlain et. al. The /data folder contains half-hourly eddy covariance files used in the analysis, the /R folder contains scripts used to process data and information theory functions, and the /vignettes folder contains the Markdown and bibiography files used to generate the manuscript.

## Installation

You can install salinity from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("samdchamberlain/salinity_study")
```

## How to Run
Load package into RStudio and open the 'manuscript.Rmd' within the /vignettes folder. Using RStudio, click the 'Knit' button at the top of the console and this manuscript, and analyses herein, will be re-created.
