---
title: Salinity study
author: Samuel D Chamberlain^1^, Elke Eichelmann^1^, Kyle S Hemes^1^, Daphne J Szutu^1^, Joseph
  G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document: default
  html_document: default
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
link-citations: yes
csl: global-change-biology.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
library(caret)
library(modelr)
library(forecast)
library(entropy)
```

\newpage

# Abstract

# Introduction

# Materials and Methods

# Results
## Salinity effects
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat6_processing.R") #wetland experiencing salinity rise
source("R/peat19_processing.R") #baseline mature wetland w/o salinity rise
```

```{r fig1, fig.height = 8, fig.width = 7}
peat6_sub <- select(peat6_daily, -starts_with("DO"), -starts_with("CO", ignore.case = F))

both_sites <- rbind(peat19_daily, peat6_sub)

a <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=mgCH4, color=site)) +
  geom_point()

b <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=gCO2, color=site)) +
  geom_point()

c <- ggplot(subset(both_sites, year > 2013)) +
  geom_line(aes(x=datetime, y=gER, color=site)) #+
  #geom_line(aes(x=datetime, y=gGEP, color=site))

d <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=Tair, color=site)) +
  geom_line()

grid.arrange(a, b, c, d, ncol=1)
```

```{r fig2, fig.height=7, fig.width=4}
#Average monthly fluxes prior to salinity peak in 2016 
MBav <- subset(peat6_monthly, year > 2012 & year < 2016) %>%
  group_by(month) %>%
  summarise(CH4 = mean(gCH4),
            sdCH4 = 1.96*sd(gCH4)/sqrt(length(gCH4)),
            NEE = mean(gNEE),
            sdNEE = 1.96*sd(gNEE)/sqrt(length(gNEE)),
            ER = mean(gER),
            sdER = 1.96*sd(gER)/sqrt(length(gER)),
            GEP = mean(gGEP),
            sdGEP = 1.96*sd(gGEP)/sqrt(length(gGEP)),
            n = length(gCH4))

#Mean conductivity on Twitchell Island without salinity rise
Tw_av <- rbind(peat19_monthly, peat6_monthly) %>%
  group_by(month) %>%
  summarise(sal = mean(mCond, na.rm=T),
            sdsal = 1.96*sd(mCond, na.rm=T)/sqrt(length(mCond)))


a <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=CH4-sdCH4, ymax=CH4+sdCH4), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=CH4)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gCH4), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gCH4), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gCH4), color="#2c7bb6") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  ylab(expression(F[CH4]~" (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="a ", vjust=1, hjust=1) +
  theme_bw() + theme(axis.title.x=element_blank())

b <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=ER-sdER, ymax=ER+sdER), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=ER)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gER), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gER), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gER), color="#2c7bb6") +
  ylab(expression("ER (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="b ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw() + theme(axis.title.x=element_blank())

c <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=GEP-sdGEP, ymax=GEP+sdGEP), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=GEP)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gGEP), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gGEP), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gGEP), color="#2c7bb6") +
  ylab(expression("GEP (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="c ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw() + theme(axis.title.x=element_blank())

d <- ggplot() +
  geom_ribbon(data=Tw_av, aes(x=month, ymin=sal-sdsal, ymax=sal+sdsal), fill="light grey") +
  geom_line(data=Tw_av, aes(x=month, y=sal)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=mCond), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=mCond), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=mCond), color="#2c7bb6") +
  ylab("Salinity")  + annotate("text", x=Inf, y=Inf, label="d ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw()

#Multiple plots aligned
g1 <- ggplotGrob(a)
g2 <- ggplotGrob(b)
g3 <- ggplotGrob(c)
g4 <- ggplotGrob(d)
g <- rbind(g1, g2, g3, g4, size="first");

grid.arrange(g)
```

```{r fig3}
ggplot(subset(peat6_yearly, year > 2011 & year < 2018), aes(x = Cond, y = tGEP, color=factor(year))) +
  geom_point() +
  stat_smooth(method="lm")
```

## Models by year for comparison
```{r anomaly_filter}
anom_filter <- function(var, window = 48*7) {
  var - ma(var, order = window)
}

test <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  map_df(anom_filter)
  
cor(test, method = "spearman", use = "complete.obs")

```

```{r pca}
subset <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD, Ts, Tw, PA, gER, Cond)

pca_test <- preProcess(subset, method = c("medianImpute"))
out_sub <- predict(pca_test, subset)

test <- prcomp(out_sub, scale. = T)

biplot(test)
```

```{r entropy_analysis}
mutual_info <- function(x, y, bins = 10) {
  
  clip_set <- na.omit(data.frame(x, y))
  
  y2d <- discretize2d(clip_set$x, clip_set$y, numBins1 = bins, numBins2 = bins)
  y_entropy <- entropy(discretize(na.omit(y), numBins = bins))
  
  mi.empirical(y2d)/y_entropy
}

mutual_info(peat6_daily$Tw, peat6_daily$mgCH4)

lag_mi <- vector("double", lags)
for (i in 1:lags) {
  x_lag <- dplyr::lag(test$Tair, n = i)
  lag_mi[[i]] <- mutual_info(x_lag, peat6_daily$mgCH4, bins = 10)
}


#calculate at multiple lags of x
lag_mi <- function(x, y, lags, bins = 10) {
  
  lag_mi <- vector("double", lags) #store lagged MI values
  
  for (i in 1:lags) {
    x_lag <- dplyr::lag(x, n = i)
    lag_mi[[i]] <- mutual_info(x_lag, y, bins)
  }
  
  plot(lag_mi, xlab = "Time lag", ylab = "MI (relative)") 
}

mutual_info(peat6_all$TA.x, peat6_all$wm, bins = 10)

# test multiple binning strategies
bin_v = seq(4, 22, by = 2)
par(mfrow=c(2, 5))

for (i in bin_v) {
  lag_mi(peat6_all$ustar, peat6_all$wm, lags = 48, bins = i)
}

#create a growing month stamp
peat6_all$month_round <- floor(peat6_all$decday/31)

#create data list by month
by_month <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(month_round) %>%
  nest()

mi_timeseries <- function(x, y, data_list, bins = 10) {

  mi_series <- vector("double", nrow(data_list)) #vector to store mutual information
  
  #iterate through indexed datasets and calculate relative MI
  for (i in 1:nrow(data_list)) {
    current_df <- data_list$data[[i]] #extract current dataframe
    x_var <- eval(substitute(x), current_df)
    y_var <- eval(substitute(y), current_df)

    mi_series[[i]] <- mutual_info(x_var, y_var, bins)
  }
  mi_series
}

GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm_gf, data_list = by_month)


#plot that needs to be dealt with later
clipped_monthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_monthly$T_mi <- t_monthly
clipped_monthly$GPP_mi <- mi_monthly

a <- ggplot(clipped_monthly) +
  geom_point(aes(x = datetime, y = ma(T_mi, 3)), color = "red") +
  geom_line(aes(x = datetime, y = ma(T_mi, 3)), color = "red") +
  geom_point(aes(x = datetime, y = ma(GPP_mi, 3)), color = "dark green") +
  geom_line(aes(x = datetime, y = ma(GPP_mi, 3)), color = "dark green") +
  ylab("Mutual information") + theme_bw() + xlab("")

b <- ggplot(clipped_monthly, aes(x=datetime, y=gCH4)) +
  geom_line() + theme_bw() + xlab("") +
  ylab("CH4 flux (g C)")

c <- ggplot(clipped_monthly, aes(x=datetime, y=gGEP)) +
  geom_line(color = "dark green") + geom_point(color = "dark green") +
  theme_bw() + xlab("") +
  ylab("GEP flux (g C)")

d <- ggplot(clipped_monthly, aes(x=datetime, y=mCond)) +
  geom_line(color = "blue") +
  ylab("Conductivity (mS)") + theme_bw()
  
  
grid.arrange(a, b, c, d, ncol=1)

```

```{r test}
#CH4sub <- peat6_all[!is.na(peat6_all$wm), ]

by_year <- peat6_daily %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(year) %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  nest()

net_model <- function(df) {
  train(mgCH4 ~ ., data = df, method = "glmnet",
                 trControl = trainControl(method = "cv"),
                 preProcess = c("center", "scale", "medianImpute", "BoxCox"),
                 tuneGrid = expand.grid(.alpha = 1, .lambda = 0.1),
                 na.action = na.pass)
}

# hh_model <- function(df) {
#   train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD,
#         data = df,
#         method = "glmnet",
#         preProcess = c("center", "scale", "medianImpute", "BoxCox"),
#         na.action = na.pass)
# }

# subset <- peat6_daily %>%
#   ungroup() %>%
#   select(mgCH4, gGEP, gER, mET, H, Tair, Tw, mVPD, u., Rain, WTD) %>%
#   na.omit()

# #caret approach
# glm_reg <- train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD, 
#                  data = CH4sub, method = "glmnet",
#                  trControl = trainControl(method = "cv"),
#                  preProcess = c("center", "scale", "medianImpute"),
#                  #tuneGrid = expand.grid(.alpha = 1, .lambda = c(0, 0.001, 0.1, 1, 5, 10, 20, 50)),
#                  na.action = na.pass)

by_year <- by_year %>%
  mutate(model = map(data, net_model),
         Imp = map(model, varImp))

impo_list <- by_year$Imp

a <- impo_list[[1]]$importance
a$year <- 2012
a$vars <- rownames(a)

b <- impo_list[[2]]$importance
b$year <- 2013
b$vars <- rownames(b)

c <- impo_list[[3]]$importance
c$year <- 2014
c$vars <- rownames(c)

d <- impo_list[[4]]$importance
d$year <- 2015
d$vars <- rownames(d)

e <- impo_list[[5]]$importance
e$year <- 2016
e$vars <- rownames(e)

f <- impo_list[[6]]$importance
f$year <- 2017
f$vars <- rownames(f)

all_now <- do.call("rbind", list(a, b, c, d, e, f))

ggplot(all_now, aes(x=year, y=Overall, color=vars)) +
  #stat_smooth(se=F) +
  geom_point(size = 2) +
  geom_line()

# ggplot(all_now, aes(x = year, y = vars)) +
#   geom_tile(aes(fill = Overall))
```

## Information theory approach


## Water chemistry

```{r fig4}
only16 <- subset(peat6_daily, year > 2015)
only16$jday <- lubridate::yday(only16$datetime)

only16$season <- ifelse(only16$jday %in% c(0:100) | only16$jday %in% c(300:365), "fallow", 
                           "growing")

ggplot(only16, aes(x=datetime)) +
  geom_line(aes(y = DOrange1), color="red") +
  geom_line(aes(y = DOdiff), color="blue")

ggplot(only16, aes(x=mgCH4, y=DOrange1, color=season)) +
  geom_point()

regr <- lm(DOrange1 ~ mgCH4, data = subset(only16, season == "growing")); summary(regr)
```

```{r fig5}
only16 <- subset(peat6_daily, year == 2016)

only16$maCH4 <- only16$gCO2 - ma(only16$gCO2, 14)
only16$maDO <- only16$DOrange1 - ma(only16$DOrange1, 14)

ggplot(only16, aes(x=maDO, y=maCH4)) +
  geom_point()
```

# Discussion

# Acknowledgements
This research was supported in part by the U.S. Department of Energy’s Office of Science and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

