---
title: "Salinity study"
author: Samuel D Chamberlain^1^, Elke Eichelmann^1^, Kyle S Hemes^1^, Daphne J Szutu^1^,
  Joseph G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
link-citations: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
fontsize: 12pt
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
library(caret)
library(modelr)
library(forecast)
library(entropy)
library(stringr)
```

<!-- \newpage -->

<!-- # Abstract -->

<!-- # Introduction -->

<!-- # Materials and Methods -->

<!-- # Results -->
## Salinity effects
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat6_processing.R") #wetland experiencing salinity rise
source("R/peat19_processing.R") #baseline mature wetland w/o salinity rise
```

```{r fig1, fig.height = 8, fig.width = 7}
peat6_sub <- select(peat6_daily, -starts_with("DO"), -starts_with("CO", ignore.case = F))

both_sites <- rbind(peat19_daily, peat6_sub)

a <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=mgCH4, color=site)) +
  geom_point(size = 0.5)

b <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=gCO2, color=site)) +
  geom_point(size = 0.5)

c <- ggplot(subset(both_sites, year > 2013)) +
  geom_line(aes(x=datetime, y=gER, color=site)) #+
  #geom_line(aes(x=datetime, y=gGEP, color=site))

d <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=Tair, color=site)) +
  geom_line()

grid.arrange(a, b, c, d, ncol=1)
```

```{r fig2, fig.height=7, fig.width=3.5}
#Average monthly fluxes prior to salinity peak in 2016 
MBav <- subset(peat6_monthly, year > 2012 & year < 2016) %>%
  group_by(month) %>%
  summarise(CH4 = mean(gCH4),
            sdCH4 = 1.96*sd(gCH4)/sqrt(length(gCH4)),
            NEE = mean(gNEE),
            sdNEE = 1.96*sd(gNEE)/sqrt(length(gNEE)),
            ER = mean(gER),
            sdER = 1.96*sd(gER)/sqrt(length(gER)),
            GEP = mean(gGEP),
            sdGEP = 1.96*sd(gGEP)/sqrt(length(gGEP)),
            n = length(gCH4))

#Mean conductivity on Twitchell Island without salinity rise
Tw_av <- rbind(peat19_monthly, peat6_monthly) %>%
  group_by(month) %>%
  summarise(cond = mean(mCond, na.rm=T),
            sd.cond = 1.96*sd(mCond, na.rm=T)/sqrt(length(mCond)))


# plot theme
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.x = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

# four panels
a <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=CH4-sdCH4, ymax=CH4+sdCH4), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=CH4)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gCH4, color=factor(year))) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  ylab(expression(F[CH4]~" (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="a ", vjust=1, hjust=1, fontface = "bold") +
  time_series + 
  theme(legend.title=element_blank(), legend.margin = margin(0, 0, 0, 0),
        legend.text = element_text(size=8), legend.position=c(.15, .75),
        legend.background = element_rect(fill = "transparent"))

b <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=ER-sdER, ymax=ER+sdER), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=ER)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gER, color=factor(year))) +
  ylab(expression("ER (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="b ", vjust=1, hjust=1,fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

c <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=GEP-sdGEP, ymax=GEP+sdGEP), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=GEP)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gGEP, color=factor(year))) +
  ylab(expression("GEP (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="c ", vjust=1, hjust=1, fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

d <- ggplot() +
  geom_ribbon(data=Tw_av, aes(x=month, ymin=cond-sd.cond, ymax=cond+sd.cond), fill="light grey") +
  geom_line(data=Tw_av, aes(x=month, y=cond)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=mCond, color=factor(year))) +
  ylab("Conductivity (mS)")  + 
  annotate("text", x=Inf, y=Inf, label="d ", vjust=1, hjust=1, fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

#Multiple plots aligned
g1 <- ggplotGrob(a)
g2 <- ggplotGrob(b)
g3 <- ggplotGrob(c)
g4 <- ggplotGrob(d)

grid.arrange(g1, g2, g3, g4, ncol = 1)
```

## Connectivity across time

```{r pca, echo = F}
subset <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD, Ts, Tw, PA, gER, Cond)

pca_test <- preProcess(subset, method = c("medianImpute"))
out_sub <- predict(pca_test, subset)

test <- prcomp(out_sub, scale. = T)

biplot(test)
```

```{r entropy_analysis_peat, fig.cap="Monthly relative mutual information (MI; top panel) shared between GEP and F~CH4~ (green line) as well as T~a~ and F~CH4~ (red line). All MI values are corrected for bias due to varied sample size using Monte Carlo reshuffling of time series such that MI = 0 represents the threshold of statistical significance (*P* < 0.05). Bottom panel displays daily mean water column conductivity (mS).", fig.width=7, fig.height=5, dpi=600}
source("R/MI_functions.R")

# # test multiple binning strategies
# bin_v = seq(4, 22, by = 2)
# par(mfrow=c(2, 5))
# 
# for (i in bin_v) {
#   a <- lag_mi(peat6_all$ustar, peat6_all$wm, lags = 48, bins = i)
#   plot(a)
# }

#create a growing month stamp
peat6_all$month <- month(peat6_all$datetime)
peat6_all$m_yr <- paste(peat6_all$year, peat6_all$month, sep = "")

#shuffle data for comparison w/ monthly gap magnitude retained
peat6_all <- peat6_all %>%
  group_by(month) %>%
  mutate(H_shuf = sample(TA.x, length(TA.x), replace = F),
         G_shuf = sample(gpp_ANNnight, length(gpp_ANNnight), replace = F),
         wm_shuf = sample(wm, length(wm), replace = F))

#create data list by month
by_month <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(m_yr) %>%
  nest()

#plot that needs to be dealt with later
clipped_monthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_monthly$date <- as.Date(clipped_monthly$datetime)
clipped_monthly$GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm, data_list = by_month)
clipped_monthly$ET_CH4_mi <- mi_timeseries(H_shuf, wm_shuf, data_list = by_month)
clipped_monthly$Tair_CH4_mi <- mi_timeseries(TA.x, wm, data_list = by_month)

# Monte Carlo significance thresholds (dont run everytime as they take a while)
#sig_gpp <- conf_series(gpp_ANNnight, wm, data_list = by_month, runs = 1000, alpha = 0.01)
# mean sig_gpp across all months at p < 0.01: 0.034

#sig_Tair <- conf_series(TA.x, wm, data_list = by_month, runs = 1000, alpha = 0.01)
# mean sig_Tair across all months at p < 0.01: 0.042

# sig_H <- conf_series(H_gf, wm, data_list = by_month, runs = 1000, alpha = 0.01)
# # mean sig_Tair across all months at p < 0.01: 0.039

# clipped_monthly$Tair_ma <- na.ma(ma(clipped_monthly$Tair_ma, 3))
# clipped_monthly$GPP_ma <- na.ma(ma(clipped_monthly$GPP_ma, 3))

# Correct MI values of bias due to sample size, now anything over zero is significant at p < 0.01
clipped_monthly$TairMI_corrected <- clipped_monthly$Tair_CH4_mi - sig_Tair
clipped_monthly$gppMI_corrected <- clipped_monthly$GPP_CH4_mi - sig_gpp

a <- ggplot(clipped_monthly) +
  geom_point(aes(x = date, y = TairMI_corrected), color = "red", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(TairMI_corrected, 3)), color = "red", size = 1) +
  geom_point(aes(x = date, y = gppMI_corrected), color = "dark green", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(gppMI_corrected, 3)), color = "dark green", size = 1) +  
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  ylab(expression(bolditalic(MI)~"(x,"~F[CH4]~")")) + theme_bw() + xlab("")

b <- ggplot(subset(peat6_daily, year > 2011 & year < 2018), aes(x=as.Date(datetime), y=Cond)) +
  geom_line(color = "blue") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  ylab("Conductivity (mS)") + xlab("Year") +
  theme_bw()
  
grid.arrange(a, b, ncol=1)
```

```{r tr_GPP_to_CH4, fig.cap="Normalized transfer entropy (***T***) from (top panel) half-hourly GEP to F~CH4~ and (bottom panel) T~a~ to F~CH4~ over the growing season during years of low salinity (2012-2013) and high salinity (2015-2016). The red line represents significance threshold (*P* < 0.05) based on Monte Carlo reshuffling of the time series. Other lines represented ***T*** at various lags of GEP or T~a~ and color demonstrates mean conductivity (mS) during the respective year. Years 2014 and 2017 are omitted from analysis because other disturbance events affected wetlands processes and confound entropy analyses." fig.height=4, fig.width=7.5, dpi=600}
source("R/TR_functions.R")
source("R/transformations.R")

#apply anomaly filter
anom_data <- peat6_all %>%
  arrange(year, time) %>%
  mutate(CH4_anom = anom_filter(wm_gf),
         TA_anom = anom_filter(TA.x),
         GPP_anom = anom_filter(gpp_ANNnight))

anom_data <- arrange(anom_data, datetime)
anom_data$CH4na_anom <- ifelse(is.na(anom_data$wm), NA, anom_data$CH4_anom) # re-insert missing values

# Focus only on high v. low salinity years without additional disturbance
# other disturbance includes the 2014 insect outbreak and 2017 flushing events
anom_data <- subset(anom_data, year %in% c(2012, 2013, 2015, 2016))

#create data list by annual growing season
by_year <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  nest()

#calculate lagged transfer entropy time series (up to 2 day lag at half-hourly timestep)
# for GPP -> CH4
tr_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 120, bins = 10)

mc_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 120, bins = 10, runs = 5, alpha = 0.05)
# for TA -> CH4
tr_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 120, bins = 10)

mc_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 120, bins = 10, runs = 5, alpha = 0.05)

#unlist data to tidy dataframe
gpp.wm_df <- tidy_list(tr_wm_gpp); gpp.wm_shuf <- tidy_list(mc_wm_gpp)
ta.wm_df <- tidy_list(tr_wm_ta); ta.wm_shuf <- tidy_list(mc_wm_ta)

#add mean annual conductivity to dataframes by year
mean_cond <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  summarise(`Cond (mS)` = mean(Cond, na.rm = T))

gpp.wm_df <- merge(gpp.wm_df, mean_cond, all.x = T) 
ta.wm_df <- merge(ta.wm_df, mean_cond, all.x = T) 

a <- ggplot() +
  geom_line(data = gpp.wm_df, aes(x = hour, y = tr, color = `Cond (mS)`)) +
  geom_line(data = gpp.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(limits = c(0, 60), breaks = c(0, 12, 24, 36, 48, 60)) +
  xlab("") + ylab(expression(bolditalic(T)~"("~GEP %->% F[CH4]~")"))

b <- ggplot() +
  geom_line(data = ta.wm_df, aes(x = hour, y = tr, color = `Cond (mS)`)) +
  geom_line(data = ta.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(limits = c(0, 60), breaks = c(0, 12, 24, 36, 48, 60)) +
  xlab("Lag (hours)") + ylab(expression(bolditalic(T)~"("~T[a] %->% F[CH4]~")"))

grid.arrange(a, b, ncol = 1)
```


```{r bias_by_samples, fig.height= 5, fig.width=6}
#use monthly from above and compare to two month chunks
# probably move to supplement

# # create a bimonthly grouping
# peat6_all$bimonth <- paste(peat6_all$year, floor((peat6_all$month+1)/2), sep="")
# 
# by_bimonth <- peat6_all %>%
#   filter(year > 2011 & year < 2018) %>%
#   group_by(bimonth) %>%
#   nest()

# #reshuffle by gap magnitude preserved
# peat6_all <- peat6_all %>%
#   group_by(bimonth) %>%
#   mutate(H_shuf = sample(TA.x, length(TA.x), replace = F),
#          G_shuf = sample(gpp_ANNnight, length(gpp_ANNnight), replace = F),
#          wm_shuf = sample(wm, length(wm), replace = F))

# #calculate bimonthly MI
# clipped_bimonthly <- subset(peat6_monthly, year > 2011 & year < 2018)
# clipped_bimonthly$bimonth <- paste(clipped_bimonthly$year, floor((clipped_bimonthly$month+1)/2), sep="")
# clipped_bimonthly <- aggregate(clipped_bimonthly, by=list(clipped_bimonthly$bimonth), FUN=mean)

clipped_monthly$T_shuf <- mi_timeseries(H_shuf, wm_shuf, data_list = by_month)
clipped_monthly$G_shuf <- mi_timeseries(G_shuf, wm_shuf, data_list = by_month)

ggplot() +
  geom_point(data = clipped_monthly, aes(x = CH4_obs, y = G_shuf)) +
  geom_point(data = clipped_monthly, aes(x = CH4_obs, y = T_shuf), color = "red") +
  xlab("Number of flux obs.") + ylab("MI (reshuffled data)")
```

```{r test, echo = F}
#CH4sub <- peat6_all[!is.na(peat6_all$wm), ]

by_year <- peat6_daily %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(year) %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  nest()

net_model <- function(df) {
  train(mgCH4 ~ ., data = df, method = "glmnet",
                 trControl = trainControl(method = "cv"),
                 preProcess = c("center", "scale", "medianImpute", "BoxCox"),
                 tuneGrid = expand.grid(.alpha = 1, .lambda = 0.1),
                 na.action = na.pass)
}

# hh_model <- function(df) {
#   train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD,
#         data = df,
#         method = "glmnet",
#         preProcess = c("center", "scale", "medianImpute", "BoxCox"),
#         na.action = na.pass)
# }

# subset <- peat6_daily %>%
#   ungroup() %>%
#   select(mgCH4, gGEP, gER, mET, H, Tair, Tw, mVPD, u., Rain, WTD) %>%
#   na.omit()

# #caret approach
# glm_reg <- train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD, 
#                  data = CH4sub, method = "glmnet",
#                  trControl = trainControl(method = "cv"),
#                  preProcess = c("center", "scale", "medianImpute"),
#                  #tuneGrid = expand.grid(.alpha = 1, .lambda = c(0, 0.001, 0.1, 1, 5, 10, 20, 50)),
#                  na.action = na.pass)

by_year <- by_year %>%
  mutate(model = map(data, net_model),
         Imp = map(model, varImp))

impo_list <- by_year$Imp

a <- impo_list[[1]]$importance
a$year <- 2012
a$vars <- rownames(a)

b <- impo_list[[2]]$importance
b$year <- 2013
b$vars <- rownames(b)

c <- impo_list[[3]]$importance
c$year <- 2014
c$vars <- rownames(c)

d <- impo_list[[4]]$importance
d$year <- 2015
d$vars <- rownames(d)

e <- impo_list[[5]]$importance
e$year <- 2016
e$vars <- rownames(e)

f <- impo_list[[6]]$importance
f$year <- 2017
f$vars <- rownames(f)

all_now <- do.call("rbind", list(a, b, c, d, e, f))

ggplot(all_now, aes(x=year, y=Overall, color=vars)) +
  #stat_smooth(se=F) +
  geom_point(size = 2) +
  geom_line()
```

<!-- ## Information theory approach -->


<!-- ## Water chemistry -->

<!-- ```{r fig4, echo = F} -->
<!-- only16 <- subset(peat6_daily, year > 2015) -->
<!-- only16$jday <- lubridate::yday(only16$datetime) -->

<!-- only16$season <- ifelse(only16$jday %in% c(0:100) | only16$jday %in% c(300:365), "fallow",  -->
<!--                            "growing") -->

<!-- # ggplot(only16, aes(x=datetime)) + -->
<!-- #   geom_line(aes(y = DOrange1), color="red") + -->
<!-- #   geom_line(aes(y = DOdiff), color="blue") -->
<!-- #  -->
<!-- # ggplot(only16, aes(x=mgCH4, y=DOrange1, color=season)) + -->
<!-- #   geom_point() -->

<!-- regr <- lm(DOrange1 ~ mgCH4, data = subset(only16, season == "growing")); summary(regr) -->
<!-- ``` -->

<!-- # Discussion -->

# Acknowledgements
This research was supported in part by the U.S. Department of Energy’s Office of Science and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

