---
title: "Salinity study"
author: Samuel D Chamberlain^1^, Elke Eichelmann^1^, Kyle S Hemes^1^, Daphne J Szutu^1^,
  Joseph G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document: default
  html_document: default
link-citations: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
fontsize: 12pt
csl: ecology-letters.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(caret)
library(forecast)
library(entropy)
library(stringr)
library(wq)
```

# Abstract

# Introduction
Peatland restoration is a commonly proposed natural climate mitigation strategy due to long-term sequestration of carbon dioxide (CO~2~) in anoxic soils [@griscom_natural_2017; @leifeld_underappreciated_2018], which also reverses large CO~2~ emissions from the drained, oxidizing peat soils [@hatala_greenhouse_2012; @knox_agricultural_2015]. From a climate mitigation perspective it is often assumed that methane (CH~4~) emissions from re-flooded landscapes are offset by carbon (C) storage over multi-century time-scales [@mitsch_wetlands_2013; @griscom_natural_2017; @leifeld_underappreciated_2018]. Despite long-term climate benefits, CH~4~ emissions should be considered when assessing the greenhouse gas (GHG) benefit of wetland restoration activities given the strength of these sustained emissions relative to CO~2~ uptake [@neubauer_moving_2015] and the observed response of surface radiative forcing to changes in global CH~4~ emissions over short (annual to seasonal) time scales [@feldman_observationally_2018]. A global network of studies, where ecosystem-scale fluxes of CO~2~ and CH~4~ can be measured *in situ* by eddy covariance [@baldocchi_fluxnet_2001], paint a less clear picture of this climate benefit, as many wetlands are net GHG sources due to CH~4~ emissions over 20 to 100 year timescales [@petrescu_uncertain_2015]. There is a strong need to understand wetland responses to disturbance and management as well if the goal is to engineer wetlands to have a limited climate footprint.

Salinization is a common disturbance affecting wetlands globally, where agricultural practices [@williams_anthropogenic_2001] and road salting [@kaushal_increased_2005] commonly effect inland freshwater systems, while coastal wetlands may be affected by drought or management-induced saltwater intrusion [@drexler_effect_2001; @barlow_saltwater_2010]. This is of particular concern in arid to semi-arid regions where evaporative demands are high and precipitation/freshwater inflow rates are low [@williams_salinisation_2002; @jolly_review_2008], and salinity levels vary with regional precipitation and water demand [@enright_salinity_2009]. Wetland salinization is expected to become more common in Mediterranean regions where water resources are scarce and aridity is predicted to increase with changing climate [@iglesias_challenges_2007; @gao_increased_2008]. The effect of changing salinity on wetlands are far reaching, affecting C cycling and export [@chambers_effect_2013; @ardon_drought_2016], nutrient cycling and biogeochemistry [@morrissey_salinity_2014; @dijk_salinization_2015], and biological community composition [@nielsen_modified_2009]. In terms of GHG exchange, if seawater (containing sulfate) is the main source of increasing salinity CH~4~ emissions will be inhibited by changing redox conditions [@poffenbarger_salinity_2011], and for this reason coastal wetlands are often the focus of efforts where restoring tidal flows can limit CH~4~ emissions from wetland complexes [@kroeger_restoring_2017]. However, increasing salinity may increase net GHG emissions due to stress-induced reductions in photosynthesis that offset perceived benefits of reduced CH~4~ emission [@neubauer_ecosystem_2013; @krauss_component_2016]. Additionally, gross ecosystem photosynthesis (GEP) provides substrate for wetland CH~4~ production [@hatala_gross_2012], so interpreting responses to salinization or other disturbance requires quantifying interactions across multiple GHG exchanges. More generally, our understanding of how wetland GHG exchange responds to disturbance and management is limited. Fluxes can vary strongly over small spatial scales due to variable soil properties [@chamberlain_soil_2018] and vegetation structure [@matthes_parsing_2014], and often exhibit substantial inter-annual variability due to disturbance and successional processes [@anderson_variation_2016; @wilson_multiyear_2016; @oikawa_evaluation_2017; @pugh_wetland_2018]. A better functional understanding wetland GHG exchange and responses to disturbance and management is required for a more robust assessment of the climate benefit of wetland restoration.

Understanding such responses can be challenging, as ecosystems are complex systems where non-linear, lagged, and feedback interactions are common [@ruddell_ecohydrologic_2009; @sturtevant_identifying_2016]. In some cases, findings from small-scale experiments may not be relevant to whole ecosystem dynamics [@schindler_dilemma_2012], which requires tools to delineate dominant or causal interactions from *in situ* observations. A number of statistical approaches common to neuroscience and econometrics, yet relatively novel to ecology, are suited to this task [@rinderer_assessing_2018], and are useful for constructing appropriate models to predict future ecosystem states [@larsen_appropriate_2016; @getz_wayne_m_making_2017]. These include non-parametric information theory metrics, such as mutual information (MI), that can identify dominant lagged, non-linear, and non-monotic relationships commonly observed in flux timeseries [@sturtevant_identifying_2016; @chamberlain_soil_2018; @knox_direct_2018]. Other metrics aimed at identifying causal relationships, such as Granger causality, have been used to identify causal relationships between GHG fluxes and the critical timescales at which they occur in forest [@detto_causality_2012] and rice ecosystems [@hatala_gross_2012]. Transfer entropy (TE), an additional information theory metric equivalent to Granger causality under Gaussian conditions [@barnett_granger_2009], has also been applied to successfully to understand functional changes in ecohydrologic systems during drought [@ruddell_ecohydrologic_2009; @ruddell_ecohydrologic_2009-1] and how C cycling responds to restoration practices [@larsen_disrupted_2017]. These statistical methods provide a means to better understand functional changes across disturbance events, which is critical to improving our understanding of how restored systems might respond to management or disturbance.

The objectives of this study were to determine the effect of drought-induced wetland salinization on CO~2~ and CH~4~ fluxes from restored wetlands in the Sacramento-San Joaquin Delta of California, USA. We measured ecosystem-scale CO~2~ and CH~4~ fluxes by eddy covariance from a restored wetland complex that experienced a substantial rise in salinity over the course of five years, and was then subsequently flushed with freshwater by land managers during the sixth year. We then used (1) information theory metrics to identify the how couplings and causal relationships between temperature, GEP, and CH~4~ fluxes change over the course of wetland salinization, and (2) used machine learning approaches to quantify the overall reduction in GEP and CH~4~ fluxes due to salinization. Understanding and predicting the response of Delta wetland GHG exchange to salinization events is important from a policy perspective, as wetland restoration in the region is funded through California's Cap-and-Trade Program [@deverel_implications_2017; @oikawa_evaluation_2017] and freshwater is projected to become an increasingly limited resource in the state's future [@vicuna_sebastian_sensitivity_2007].

# Materials and Methods
## Site Description and Flux Measurements
We measured ecosystem-scale fluxes of CO~2~, CH~4~, water (H~2~O), and energy for seven years from a restored wetlands on Sherman Island in the Sacramento-San Joaquin Delta of California, USA (Ameriflux site Us-Myb; N 38.0498, W 121.7650). The wetland was restored in 2010 on a former pasture, and our measurements began at the point of initial restoration and re-flooding. The site is predominantly vegetated with cattails (*Typha* spp.) and some tules (*Schoenoplectus acutus*), and does not experience tidal changes in salinity experienced by the adjacent Sacramento and San Joaquin rivers. The wetland's water is closed off from river sources, though managers are able to control the inflow and outflow of water to the wetland from the main river channels. We also compared findings from this site to a mature wetland reference site (Ameriflux site Us-Tw1; N 38.1074, W 121.6469) restored in 1999 that did not experience an increase in salinity across the same time period. More description of this mature reference site can be found in @miller_plant_2010 and @eichelmann_effect_2018.

Fluxes were measured continuously by eddy covariance, where fluxes were intergrated over half-hour periods from the covariance of vertical wind velocity using a 3-dimensional sonic anemometer (Gill WindMaster; Gill Instruments Ltd, Lymington, Hampshire, England) and gas concentrations (CO~2~, CH~4~, and H~2~O) using open-path infrared gas analyzers (LI-7500 & LI-7700; LI-COR, Lincoln, NE, USA) measured at 20 Hz frequency. Standardized flux corrections and quality control were applied, including high-frequency data despiking, 2-D coordinate rotations, density corrections, and site-specific friction velocity (*u$^\ast$*) filtering. Data gaps in all fluxes were filled using artificial neural networks (ANNs), and net ecosystem exchange of CO~2~ (NEE) was partitioned into ER and GEP by training an ANN on nighttime CO~2~ fluxes and using this ANN to predict ER at all times. GEP was then estimated as the difference between NEE and ER. Site instrumention and descriptions are described in greater detail in [@eichelmann_effect_2018], quality control procedures are described in more detail in @knox_agricultural_2015 and @chamberlain_evaluation_2017, and gap-filling and partitioning @baldocchi_does_2015. The partitioning methods perform well against indenpendent soil respiration verfication methods [@oikawa_revisiting_2017]. We also measured air temperature, conductivity, among a host of other auxillary variables described in @eichelmann_effect_2018. We measured water conductivity using stainless-steel electrodes (CS547A-L; Campbell Scientific, Logan, UT, USA) and air temperature using shielded and aspirated thermistor and capacitance sensors (HMP45; Viasala, Vantaa, Finland).

## Information Theory Methods
We used *MI* to identify coupling strengths and *TE* to identify causal interactions between variables and the time scales at which they occur. Mutual information is used to identify the amount of information shared between two variables *X* and *Y*, or the reduction of uncertainty given knowledge of another variable, and can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies of each variable, and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. Marginal entropies are calculated from the probability distributions of a variable *Y* as follows:
$$H_Y=-\sum_{y_t}p(y_t)log_2p(y_t)$$
And joint entropies are calculated from the probability distributions of variables *X* and *Y* as follows:
$$H_{X,Y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. We calculated *MI* for one month blocks across the entire flux time series to quantify the change in couplings between flux variables across the salinity disturbance event. For each month, continuous variables were discretized to 10 bins of equivalent length across the range of each variable. Mutual information was then calculated for each variable pair of interest (*Ta*,*F~CH4~* & *GEP*,*F~CH4~*) and a significance threshold was determined via 100 Monte Carlo reshufflings of the independent variable (*Ta* or *GEP*). Because the reshuffled values (*MI'*) varies with the number of CH~4~ flux observations per month (see Supporting Information) we present *MI-MI'*, similar to @larsen_disrupted_2017, such that any values greater than zero are signficant couplings.

While *MI* identifies coupling strength [@sturtevant_identifying_2016; @chamberlain_soil_2018] between variables it does not ascribe directionality to the relationship. Transfer entropy quantifies directional information transfer, or more technically, the extent to which knowledge of a variables *X* reduced the uncertainty of the present state of *Y* given knowledge of the history of *Y* itself. If a value of *TE* is above some signficance threshold (described below) it is deduced that information flows from *X* to the current state of *Y* [@ruddell_ecohydrologic_2009]. Transfer entropy can also be described by joint and marginal entropies incorporating time lags as follows [@knuth_revealing_2013; @ruddell_ecohydrologic_2009]:
$$T(X>Y, \tau) = H(X_{t-\tau}, Y_{t-1}) + H(Y_{t}, Y_{t-1}) - H(Y_{t-1}) + H(X_{t-\tau}, Y_{t}, Y_{t-1})$$
In the above formulation, we assume the immediate history of *Y* (t-1; i.e. the preceding half-hour in eddy flux time series) contributes the most information to the current state of *Y* and we calculate *TE* across multiple lags of *X* to identify the critical timescale of interactions between variables [@ruddell_ecohydrologic_2009; @larsen_disrupted_2017]. A more in depth description of applying *TE* and *MI* to ecosystem-atmosphere flux time series can be found in @ruddell_ecohydrologic_2009, and we follow the basic protocols outlined in this work. This includes building discrete probability distributions using 10 bins for each variable as described above, transforming variables prior to *TE* calculations using a 5-day anomaly filter, and again establishing significance thresholds using Monte Carlo reshuffling of the flux time series. We calculate *TE* for information flow from *Ta* to *F~CH4~* and *GEP* to *F~CH4~*, and did so across multiple lags of the independent variable up to 36 hours to understand the timescale of interactions. Similarly, we calculated significance thresholds across each lag and due to computational demands used 25 reshufflings, rather than 100, when calculating *TE*. All information theory calculations were conducted using R 3.3.2 [@r_citation] with the 'entropy' package and home built functions to calculate *TE* and significance thresholds available online with this manuscript at https://github.com/samdchamberlain/salinity_study.

## Determining Salinity Impacts on Annual GHG Fluxes
To quantify the reduction in *GEP* and *F~CH4~* due to salinity effects, we trained a random forest model on fluxes during years prior to the salinity rise when the wetland was fully vegetated (2012-2014) using common meteorological variables and drivers of each flux. For *GEP*, this included *Ta*, photosynthetically active radiation (*PAR*), vapor pressure deficit (*VPD*), air pressure (*PA*), friction velocity (*u$^\ast$*), and water table depth (*WTD*), and for CH~4~ flux this included *GEP*, *Ta*, *VPD*, *PA*, *u$^\ast$*, and *WTD*. Random forests are an ensemble tree-based method where numerous trees are constructed using bootstrap samples from the training dataset, and the trees are averaged to generate a final prediction. For each tree, a random subset of the predictors (*m~try~*) are chosen at each tree split to reduce overfitting [@breiman_random_2001]. We used an ensemble of 500 trees and the *m~try~* value (ranging from 2-4) was chosen by using the model formulation with the lowest root mean squared error following five-fold cross-validation [@kuhn_applied_2013]. Any missing values in the predictor data were imputed using medians. We then used the tuned random forest model to generate predictions of CH~4~ flux and *GEP* during the high salinity years (2015-2016) to quantify the overall effect of salinity on these fluxes. We were careful to only include predicitor variables that were not influenced by salinity in the model tuning. The random forest was fit using the 'caret' [@caret] and 'randomForest' [@randomForest] packages. All additional data processing, statistical analysis, and visualization were conducted using the 'tidyverse' [@tidyverse], 'zoo' [@zoo], 'gridExtra' [@gridExtra], 'lubridate' [@lubridate], and 'scales' [@scales] packages. Annual GHG budgets were calculated using the 45x 100-year sustained-flux global warming potential for CH~4~ presented in @neubauer_moving_2015. This manuscript is reproducible via R Markdown, and its code can be found at https://github.com/samdchamberlain/salinity_study.

# Results
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat6_processing.R") #wetland experiencing salinity rise
source("R/peat19_processing.R") #baseline mature wetland w/o salinity rise
```

We observed a substantial increase in salinity over the course of our observation period. Salinity levels were relatively stable from 2012 through 2013, began to rise through 2014, peaked in 2017, and reduced rapidly in 2017 when site managers flushed the wetland system with freshwater (Fig. 1). Substantial interannual variability was observed in GEP and ER fluxes, where the first year post-restoration (2012) monthly fluxes were very large in magnitude, followed by a low flux year in 2013. Methane fluxes were more stable but also reduced in size (Fig. 1). In 2015 when the salinity began to rise rapidly, all fluxes were mid-range compared across all year. In 2016 when salinity peaked, all late season fluxes were low compared to previous years (Fig. 1). Fluxes remained low into 2017, and we observed a strong increase in late season fluxes coincident with freshwater flushing that rapidly reduced salinity (Fig. 1).

Coupling between driver variables (*GEP* and *Ta*) and *F~CH4~* were similar in the pre-disturbance period (2012 & 2014) where maximum *MI* was observed during the growing season and were of similar magnitude for *MI~Ta,FCH4~* and *MI~GEP,FCH4~* (Fig. 2). During 2014, *MI* did not display a similar seasonality, presumabely due to a large-scale insect outbreak that affected the wetland, though at late season increase in *MI* was still observed for *MI~Ta,FCH4~* and *MI~GEP,FCH4~*. During 2015 and 2016, the peak of the salinity disturbance, we observed a strong decoupling between *GEP* and *F~CH4~*, as well as a strengthening in the coupling between *Ta* and *F~CH4~* compared with the previous years (Fig. 2). The decoupling of *GEP* and *F~CH4~* across 2015 and 2016 is not driven by a crash in the *GEP* flux magnitude, as *GEP* magnitudes within the season were similar to previous years (Fig. 1). During the managed flushing period of 2017 when salinity levels dropped steeply, couplings between *GEP* and *F~CH4~* began to recover but were highly variable throughout the season as salinity changed (Fig. 2). These coupling changes are somewhat observable in monthly Kendall's tau, but the general patterns are obscured by large negative coeffcients observed during winter periods (Supporting Information; Fig. S2).

```{r fig1, fig.width=3, fig.height=7, dpi=600}
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=10))

vegetated_monthly <- subset(peat6_monthly, year > 2011)
vegetated_monthly$Year <- as.factor(vegetated_monthly$year)

a <- ggplot(vegetated_monthly, aes(x=month, y=gCH4, color=Year)) +
  geom_line() +
  scale_x_continuous(breaks = seq(2, 12, by=2)) +
  ylab(expression(F[CH4]~" "~"("~g~C~m^{-2}~d^{-1}~")")) +
  time_series + theme(legend.position = "none")

b <- ggplot(vegetated_monthly, aes(x=month, y=gGEP, color=Year)) +
  geom_line() +
  scale_x_continuous(breaks = seq(2, 12, by=2)) +
  ylab(expression("GEP "~"("~g~C~m^{-2}~d^{-1}~")")) +
  time_series + theme(legend.position = "none")

c <- ggplot(vegetated_monthly, aes(x=month, y=gER, color=Year)) +
  geom_line() +
  scale_x_continuous(breaks = seq(2, 12, by=2)) +
  ylab(expression("ER "~"("~g~C~m^{-2}~d^{-1}~")")) +
  time_series +
  theme(legend.margin = margin(0, 0, 0, 0),
        legend.position=c(.14, .62), legend.key.size =  unit(0.14, "in"),
        legend.text = element_text(size=8), legend.title = element_text(size=9, face = "bold")) 

d <- ggplot(vegetated_monthly, aes(x=month, y=mSal, color=Year)) +
  geom_line() +
  scale_x_continuous(breaks = seq(2, 12, by=2)) +
  ylab("Salinty (psu)") + xlab("Month") +
  theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.y = element_text(size=10), legend.position = "none")

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); 
iC$widths[2:5] <- as.list(maxWidth); iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

```{r fig2, fig.cap="Monthly relative mutual information (*MI*; top panel) shared between GEP and F~CH4~ (green line) as well as T~a~ and F~CH4~ (red line) at the wetland that experienced salinization. All *MI* values are corrected using 100 Monte Carlo reshufflings of time series, such that *MI* > 0 are statistically significant at *P* < 0.05. The lines represent 3 month running means. Bottom panel displays daily mean water column salinity (psu).", fig.width=7, fig.height=5, dpi=600}
source("R/MI_functions.R")
source("R/signficance_tests.R")

## Processing for high salinity wetland
#create a growing month stamp
peat6_all$month <- month(peat6_all$datetime)
peat6_all$m_yr <- paste(peat6_all$year, peat6_all$month, sep = "")

#create data list by month
by_month <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(m_yr) %>%
  nest()

#plot that needs to be dealt with later
clipped_monthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_monthly$date <- as.Date(clipped_monthly$datetime)
clipped_monthly$GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm, data_list = by_month)
clipped_monthly$Tair_CH4_mi <- mi_timeseries(TA.x, wm, data_list = by_month)

# Monte Carlo significance thresholds (dont run everytime as they take a while)
sig_gpp <- conf_series(gpp_ANNnight, wm, data_list = by_month, runs = 100, alpha = 0.05)
sig_Tair <- conf_series(TA.x, wm, data_list = by_month, runs = 100, alpha = 0.05)

# Correct MI values of bias due to sample size, now anything over zero is significant at p < 0.01
clipped_monthly$TairMI_corrected <- clipped_monthly$Tair_CH4_mi - sig_Tair
clipped_monthly$gppMI_corrected <- clipped_monthly$GPP_CH4_mi - sig_gpp

a <- ggplot(clipped_monthly) +
  geom_point(aes(x = date, y = TairMI_corrected), color = "red", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(TairMI_corrected, 3)), color = "red", size = 1) +
  geom_point(aes(x = date, y = gppMI_corrected), color = "dark green", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(gppMI_corrected, 3)), color = "dark green", size = 1) +  
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  scale_y_continuous(limits = c(0, 0.3)) +
  ylab(expression(bolditalic(MI)~"(x,"~F[CH4]~")")) + xlab("") + ggtitle("Rising salinity wetland") +
  theme_bw()

b <- ggplot(subset(peat6_daily, year > 2011 & year < 2018), aes(x=as.Date(datetime), y=Sal)) +
  geom_line(color = "blue") +
  scale_y_continuous(limits = c(0, 7)) +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  ylab("Salinity (psu)") + xlab("Year") +
  theme_bw()

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); 

grid.arrange(iA, iB, ncol=1)
```

This change in coupling dynamics across the salinity disturbance gradient stands in contrast to the relatively stable couplings between *F~CH4~* and either *Ta* or *GEP* at a nearby wetland complex that did not experience an increase in salinity across the same time period (Fig. 3). This wetland experienced some watertable drawdowns in 2013 and 2014 that are discussed in more detail in @sturtevant_identifying_2016, but regardless, we observed couplings of similar magnitude for *MI~Ta,FCH4~* and *MI~GEP,FCH4~* across all years (Fig. 3). Most importantly, we did no observe a similar decoupling of *F~CH4~* and *GEP* during 2015 and 2016 when salinity levels did not increase at this wetland site. Flux magnitudes were often similar magnitude or larger at this wetland complex compared with the site that experienced salinization, particularly during the high salinity periods when *GEP* decreased at wetland experiencing salinization (Supporting Information; Fig. S3).

```{r fig3, fig.cap="Monthly relative mutual information (*MI*; top panel) shared between GEP and F~CH4~ (green line) as well as T~a~ and F~CH4~ (red line) at the wetland that experienced no increase in salinity. All *MI* values are corrected using 100 Monte Carlo reshufflings of time series, such that *MI* > 0 are statistically significant at *P* < 0.05. Only positive *MI* are shown, and the lines represent 3 month running means. Bottom panel displays daily mean water column salinity (psu).", fig.width=7, fig.height=5, dpi=600}
source("R/MI_functions.R")
source("R/signficance_tests.R")

#create a growing month stamp
peat19_all$month <- month(peat19_all$datetime)
peat19_all$m_yr <- paste(peat19_all$year, peat19_all$month, sep = "")

#create data list by month
by_month <- peat19_all %>%
  filter(year > 2012 & year < 2018) %>%
  group_by(m_yr) %>%
  nest()

#plot that needs to be dealt with later
clipped_monthly <- subset(peat19_monthly, year > 2012 & year < 2018)
clipped_monthly$date <- as.Date(clipped_monthly$datetime)
clipped_monthly$GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm, data_list = by_month)
clipped_monthly$Tair_CH4_mi <- mi_timeseries(TA.x, wm, data_list = by_month)

# Monte Carlo significance thresholds (dont run everytime as they take a while)
sig_gpp <- conf_series(gpp_ANNnight, wm, data_list = by_month, runs = 100, alpha = 0.05)
sig_Tair <- conf_series(TA.x, wm, data_list = by_month, runs = 100, alpha = 0.05)

# clipped_monthly$Tair_ma <- na.ma(ma(clipped_monthly$Tair_ma, 3))
# clipped_monthly$GPP_ma <- na.ma(ma(clipped_monthly$GPP_ma, 3))

# Correct MI values of bias due to sample size, now anything over zero is significant at p < 0.01
clipped_monthly$TairMI_corrected <- clipped_monthly$Tair_CH4_mi - sig_Tair
clipped_monthly$gppMI_corrected <- clipped_monthly$GPP_CH4_mi - sig_gpp

a <- ggplot(clipped_monthly) +
  geom_point(aes(x = date, y = TairMI_corrected), color = "red", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(TairMI_corrected, 3)), color = "red", size = 1) +
  geom_point(aes(x = date, y = gppMI_corrected), color = "dark green", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(gppMI_corrected, 3)), color = "dark green", size = 1) +  
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  scale_y_continuous(limits = c(0, 0.3)) +
  ylab(expression(bolditalic(MI)~"(x,"~F[CH4]~")")) + xlab("") + ggtitle("Baseline wetland") +
  theme_bw()

b <- ggplot(subset(peat19_daily, year > 2012 & year < 2018), aes(x=as.Date(datetime), y=Sal)) +
  geom_line(color = "blue") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  scale_y_continuous(limits = c(0, 7)) +
  ylab("Salinity (psu)") + xlab("Year") +
  theme_bw()
  
grid.arrange(a, b, ncol=1)
```

We observe a similar transition in the hierarchy of coupling between the low salinity period (2012-2013) and the high salinity period (2015-2016) in the transfer of information from *GEP* or *Ta* to *F~CH4~* (Fig. 4). During the low salinity period, we observed similar magnitude information transfers from *GEP* to *F~CH4~* and from *Ta* to *F~CH4~*. Here, significant information transfers from *GEP* to *F~CH4~* occurred between 0-12 hour time lags and over the same lag period from the previous day (Fig. 4). These transfers were still observable during 2015, but by 2016 during peak salinity, only weak significant information transfers from *GEP* to *F~CH4~* were observed at 0-3 hour lags (Fig. 4). We observed a very different dynamic information moving from *Ta* to *F~CH4~* that mirrors our *MI* observations, where *TE*(*Ta* > *F~CH4~*) was of similar magnitude to *TE*(*GEP* > *F~CH4~*) during the low salinity years (2012 & 2013), and these transfers from *Ta* to *F~CH4~* strengthened during the high salinity periods in 2015 and 2016 (Fig. 4). Under most conditions, *Ta* at time lags up to ~30 hours effected current CH~4~ flux rates (Fig. 4). Transfer entropy from 2014 and 2017 was not evaluated in this analysis because insect outbreakand flushing disturbances also effected the wetlands and obscure our ability to evaluate the effect of salinization alone. We observed no significant transfer of information from *F~CH4~* to *Ta* (Supporting Information; Fig. S4), as expected since biological trace gas fluxes should not affect regional meteorologic conditions.

```{r fig4, fig.cap="Normalized transfer entropy (***TE***) from (top panel) half-hourly GEP to F~CH4~ and (bottom panel) T~a~ to F~CH4~ over the growing season during years of low salinity (2012-2013) and high salinity (2015-2016). The red line represents significance threshold (*P* < 0.05) based on Monte Carlo reshuffling of the time series. Other lines represented ***TE*** at various lags of GEP or T~a~ and color demonstrates mean salinity (psu) during the respective year. Years 2014 and 2017 are omitted from analysis because other disturbance events affected wetlands processes and confound analyses.", fig.height=4, fig.width=7.5, dpi=600}
source("R/TR_functions.R")
source("R/transformations.R")

#apply anomaly filter
anom_data <- peat6_all %>%
  arrange(year, time) %>%
  mutate(CH4_anom = anom_filter(wm_gf),
         TA_anom = anom_filter(TA.x),
         GPP_anom = anom_filter(gpp_ANNnight))

anom_data <- arrange(anom_data, datetime)
anom_data$CH4na_anom <- ifelse(is.na(anom_data$wm), NA, anom_data$CH4_anom) # re-insert missing values

# Focus only on high v. low salinity years without additional disturbance
# other disturbance includes the 2014 insect outbreak and 2017 flushing events
anom_data <- subset(anom_data, year %in% c(2012, 2013, 2015, 2016))

#create data list by annual growing season
by_year <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  nest()

#calculate lagged transfer entropy time series (up to 2 day lag at half-hourly timestep)
# for GPP -> CH4
tr_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 72, bins = 10, normalize = T)

mc_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 72, bins = 10, runs = 5, alpha = 0.05, normalize = T)
# for TA -> CH4
tr_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 72, bins = 10, normalize = T)

mc_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 72, bins = 10, runs = 5, alpha = 0.05, normalize = T)

#unlist data to tidy dataframe
gpp.wm_df <- tidy_list(tr_wm_gpp); gpp.wm_shuf <- tidy_list(mc_wm_gpp)
ta.wm_df <- tidy_list(tr_wm_ta); ta.wm_shuf <- tidy_list(mc_wm_ta)

#add mean annual conductivity to dataframes by year
mean_sal <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  summarise(`Sal. (psu)` = mean(sal, na.rm = T))

gpp.wm_df <- merge(gpp.wm_df, mean_sal, all.x = T) 
ta.wm_df <- merge(ta.wm_df, mean_sal, all.x = T) 

a <- ggplot() +
  geom_line(data = gpp.wm_df, aes(x = hour, y = tr, color = `Sal. (psu)`)) +
  geom_line(data = gpp.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = seq(0, 48, by = 6)) +
  scale_y_continuous(limits = c(0.005, 0.04)) +
  xlab("") + ylab(expression(bolditalic(TE)~"("~GEP %->% F[CH4]~")")) +
  theme(legend.margin = margin(0, 0, 0, 0),
        legend.position=c(.82, .7), legend.key.size =  unit(0.1, "in"),
        legend.text = element_text(size=8), legend.title = element_text(size=8.5, face = "bold"),
        legend.background = element_rect(fill = "transparent"))

b <- ggplot() +
  geom_line(data = ta.wm_df, aes(x = hour, y = tr, color = `Sal. (psu)`)) +
  geom_line(data = ta.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_y_continuous(limits = c(0.005, 0.04)) +
  scale_x_continuous(breaks = seq(0, 48, by = 6)) +
  xlab("Lag (hours)") + ylab(expression(bolditalic(TE)~"("~T[a] %->% F[CH4]~")")) +
  theme(legend.position = "none")

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); 

grid.arrange(iA, iB, ncol=1)
```

```{r pre_salinity_models, fig.cap="", fig.height=6, fig.width=3.5, dpi=600}
# train nn model on pre-salinity rise years and use to calculate reduction in component fluxes 
pre_rise <- peat6_daily %>%
  filter(year > 2011 & year < 2015) %>%
  select(mgCH4, gGEP, gER, mET, PAR, H, Tair, Tw, Ts, mVPD, PA, u., Rain, WTD)

set.seed(1000)
# Random forest model for GEP trained pre-salinity rise
pre_GEP <- train(gGEP ~ Tair + PAR + mVPD + PA + u. + WTD, data = pre_rise,
                      method = "rf",
                      trControl = trainControl(method = "cv", number = 5),
                      preProcess = c("medianImpute"),
                      tuneGrid = expand.grid(.mtry = c(2, 3, 4)),
                      na.action = na.pass)

set.seed(2000)
# Random forest model for FCH4 trained pre-salinity rise
pre_CH4 <- train(mgCH4 ~ gGEP + Tair + mVPD + PA + u. + WTD, data = pre_rise,
                      method = "rf",
                      trControl = trainControl(method = "cv", number = 5),
                      preProcess = c("medianImpute"),
                      ntree = 1000,
                      tuneGrid = expand.grid(.mtry = c(2, 3, 4)),
                      na.action = na.pass)

# # generate predictions for high salinity years for GEP and FCH4
# pre_rise$GEPpred <- predict.train(pre_GEP, pre_rise, na.action = na.pass)
# pre_rise$CH4pred <- predict.train(pre_CH4, pre_rise, na.action = na.pass)
# 
# # save ob v. pred summary statistics
# GEPregr <- summary(lm(gGEP ~ GEPpred, data = pre_rise))
# CH4regr <- summary(lm(mgCH4 ~ CH4pred, data = pre_rise))
# 
# a <- ggplot(pre_rise, aes(x = CH4pred, y = mgCH4)) +
#   geom_point(alpha = 0.25) +
#   geom_abline(intercept = 0, slope = 1) +
#   xlab(expression(F[CH4-Pred.]~" (mg C"~m^{-2}~d^{-1}~")")) +
#   ylab(expression(F[CH4-Obs.]~" (mg C"~m^{-2}~d^{-1}~")")) +
#   ggtitle("Pre-salinity rise (2012-2014)")
# 
# b <- ggplot(pre_rise, aes(x = GEPpred*-1, y = gGEP*-1)) +
#   geom_point(alpha = 0.25) +
#   geom_abline(intercept = 0, slope = 1) +
#   ylab(expression(GEP[Obs.]~"("~g~C~m^{-2}~d^{-1}~")")) +
#   xlab(expression(GEP[Pred.]~"("~g~C~m^{-2}~d^{-1}~")"))
# 
# grid.arrange(a, b, ncol = 1)
```

```{r fig5, fig.width=7.5, fig.height=4}
# extract data to make predictions on the salinity rise years
rise_years <- peat6_daily %>%
  filter(year > 2014 & year < 2017) %>%
  select(datetime, year, mgCH4, gGEP, gER, mET, PAR, H, Tair, Tw, Ts, mVPD, PA, u., Rain, WTD)

# generate predictions for high salinity years for GEP and FCH4
rise_years$GEPpred <- predict.train(pre_GEP, rise_years, na.action = na.pass)
rise_years$CH4pred <- predict.train(pre_CH4, rise_years, na.action = na.pass)

#plotting all the measures
a <- ggplot(rise_years) +
  geom_line(aes(x=datetime, y=mgCH4)) +
  geom_line(aes(x=datetime, y=CH4pred), color = "red") +
  ylab(expression(F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) + xlab("") +
  ggtitle("High salinity period (2015-2016)") +
  theme_classic()

b <- ggplot(rise_years) +
  geom_line(aes(x=datetime, y=gGEP*-1)) +
  geom_line(aes(x=datetime, y=GEPpred*-1), color = "red") +
  ylab(expression("GEP "~"("~g~C~m^{-2}~d^{-1}~")")) + xlab("Date") +
  theme_classic()

grid.arrange(a, b, ncol = 1)
```

```{r flux_reduction_estimates, echo=FALSE}
#calculate annual budgets for RF predictions and observations during salinity years
annual_budgets <- rise_years %>%
  group_by(year) %>%
  summarise(CH4yr = mean(mgCH4)*365/1000,
            CH4yr_rf = mean(CH4pred)*365/1000,
            GEPyr = mean(gGEP)*365,
            GEPyr_rf = mean(GEPpred)*365) %>%
  mutate(CH4pc = (CH4yr - CH4yr_rf)*100/CH4yr,
         GEPpc = (GEPyr - GEPyr_rf)*100/GEPyr)
```

Finally, we trained random forest models on the years of data prior to the large changes in *F~CH4~* coupling during 2015 and 2016 to estimate to the overall reduction in *GEP* and *F~CH4~* during those years. As stated previously, both models were fit using 5-fold cross-validation where the random forest *m~try~* was chosen in the process. During cross-validation, the final *GEP* model explained `r max(pre_GEP$results$Rsquared)*100`% of data variance (*r^2^*; *m~try~* = 4), and the final *F~CH4~* model explained `r max(pre_CH4$results$Rsquared)*100`% of the variance (*r^2^*; *m~try~* = 3).

Model predictions during the high salinity period demonstrate that CH~4~ fluxes were not substantially lower than expected if the impact of salinity was not taken into account (Fig. 5). During 2015, annual CH~4~ flux observations were only `r annual_budgets$CH4pc[annual_budgets$year == 2015]`% below the model predictions, though during 2016, observed annual budgets were `r annual_budgets$CH4pc[annual_budgets$year == 2016]`% below the model predictions. A much larger effect was observed for *GEP*, where observed *GEP* was substantially below the random forest predictions not accounting for salinity effects, particularly during the growing season of 2016 when salinity levels were highest (Fig. 5). During 2015, observed annual *GEP* budgets were `r annual_budgets$GEPpc[annual_budgets$year == 2015]`% lower than predictions, and during 2016 annual *GEP* observations were `r annual_budgets$GEPpc[annual_budgets$year == 2016]`% lower than predictions assuming no effect of salinity. This modeling exercise demonstrates the disproportionate effect of this form of wetland salinization on *GEP* rather than CH~4~ emissions.

# Discussion
Our analysis suggests that salinization affecting restored wetlands in the Sacramento-San Joaquin Delta disproportionately affects wetland *GEP* compared to CH~4~ fluxes (Fig. 5), which leads to an overall increase in wetland GHG emissions during the period of peak salinity in 2016 [@chamberlain_soil_2018]. The peak salinity this wetland experienced was well within the range where suppression of CH~4~ emissions is expected in the presence of sulphate in seawater, as observed by eddy covariance [@holm_ecosystem_2016] and chamber studies [@poffenbarger_salinity_2011], suggesting the the driver of wetland salinization was enhanced evaporation over the course of the drought rather than saltwater intrusion from the nearby estruary. These findings suggest that restored wetland GHG emissions may increase with drought-induced salinization in semi-arid to arid climates due to small decreases in CH~4~ emissions relative to the larger reductions in CO~2~ uptake if sulphate ions are not concentrated in the process.

Though we observed relatively small decreases in CH~4~ emissions during the salinization period (Fig. 4), we observed strong changes in connectivity strength (Fig. 2) and information transfer (Fig. 4) between driving processes (*Ta* and *GEP*) and CH~4~ fluxes across the disturbance period. During the high salinity period (2015-2016), couplings (*MI*) and information transfer (*TE*) from *Ta* to *F~CH4~* strengthened (Fig. 2 and 4, respectively), and coupling and information transfer from *GEP* to *F~CH4~* declined (Fig. 2 and 4, respectively). These observations, taken with the findings of small reductions in daily to annual *F~CH4~* (Fig. 5), suggest that either (1) microbial utilization of other weltand C substrates compenstates for a reduction in utilization of recent photosynthates, or (2) recent photosynthates are not an important driver of CH~4~ production and emissions within these wetland systems. This stands in constrast to previous eddy covariance studies conducted on rice paddies in the Sacramento-San Joaquin Delta, where it was found that *GEP* was the dominant driver of CH~4~ fluxes compared to temperature [@hatala_gross_2012; @knox_biophysical_2016]. Instead we observe similar coupling strengths between *F~CH4~* and *GEP* or *Ta* at the undisturbed wetland (Fig. 3) and increased temperature dependence during disturbance periods as described previously. The contrast between these findings and those from rice systems in the same region suggests that dominant biogeochemical processes can vary widely across ecosystem types and caution should be taken in extrapolating theory from one system to another.

Section on merits of information theory and causual inference.... To be to mention spurious TE here and importance of using biological knowledge and other more sophisticated approaches

Final conclusion wrap up...



# Acknowledgements
This research was supported in part by the U.S. Department of Energy’s Office of Science and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

