---
title: "Salinity study"
author: Samuel D Chamberlain^1^, Elke Eichelmann^1^, Kyle S Hemes^1^, Daphne J Szutu^1^,
  Joseph G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
link-citations: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
fontsize: 12pt
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
library(caret)
library(modelr)
library(forecast)
library(entropy)
library(stringr)
```

# Abstract

# Introduction
Restoration of drained peatlands to wetlands is an often proposed natural climate mitigation strategy due to long-term sequestration of carbon dioxide (CO~2~) in anoxic soils [@griscom_natural_2017; @leifeld_underappreciated_2018], which also reverses large CO~2~ emissions from the drained peat soils exposed to oxygen [@hatala_greenhouse_2012; @knox_agricultural_2015]. From a climate mitigation perspective it is often assumed that methane (CH~4~) emissions from re-flooded landscapes are offset by carbon (C) storage over multi-century time-scales [@mitsch_wetlands_2013; @griscom_natural_2017; @leifeld_underappreciated_2018]. Despite long-term climate benefits, CH~4~ emissions should be considered when assessing the greenhouse gas (GHG) benefit of wetland restoration activities given the strength of these sustained emissions relative to CO~2~ uptake [@neubauer_moving_2015] and the observed response of surface radiative forcing to changes in global CH~4~ emissions over short (annual to seasonal) time scales [@feldman_observationally_2018]. A global network of studies, where ecosystem-scale fluxes of CO~2~ and CH~4~ can be measured *in situ* by eddy covariance [@baldocchi_fluxnet_2001], paint a less clear picture of this climate benefit, as many wetlands are net GHG sources due to CH~4~ emissions over 20 to 100 year timescales [@petrescu_uncertain_2015]. There is a strong need to understand wetland responses to disturbance and management as well if the goal is to engineer wetlands to have a limited climate footprint.

Salinization is a common disturbance affecting wetlands globally, where agricultural practices [@williams_anthropogenic_2001] and road salting [@kaushal_increased_2005] commonly effect inland freshwater systems, while coastal wetlands may be affected by drought or management-induced saltwater intrusion [@drexler_effect_2001; @barlow_saltwater_2010]. This is of particular concern in arid to semi-arid regions where evaporative demands are high and precipitation/freshwater inflow rates are low [@williams_salinisation_2002; @jolly_review_2008], and salinity levels vary with regional precipitation and water demand [@enright_salinity_2009]. Wetland salinization is expected to become more common in Mediterranean regions where water resources are scarce and aridity is predicted to increase with changing climate [@iglesias_challenges_2007; @gao_increased_2008]. The effect of changing salinity on wetlands are far reaching, affecting C cycling and export [@chambers_effect_2013; @ardon_drought_2016], nutrient cycling and biogeochemistry [@morrissey_salinity_2014; @dijk_salinization_2015], and biological community composition [@nielsen_modified_2009]. In terms of GHG exchange, if seawater (containing sulfate) is the main source of increasing salinity CH~4~ emissions will be inhibited by changing redox conditions [@poffenbarger_salinity_2011], and for this reason coastal wetlands are often the focus of efforts where restoring tidal flows can limit CH~4~ emissions from wetland complexes [@kroeger_restoring_2017]. However, increasing salinity may increase net GHG emissions due to stress-induced reductions in photosynthesis that offset perceived benefits of reduced CH~4~ emission [@neubauer_ecosystem_2013; @krauss_component_2016]. Additionally, gross ecosystem photosynthesis (GEP) provides substrate for wetland CH~4~ production [@hatala_gross_2012], so interpreting responses to salinization or other disturbance requires quantifying interactions across multiple GHG exchanges. More generally, our understanding of how wetland GHG exchange responds to disturbance and management is limited. Fluxes can vary strongly over small spatial scales due to variable soil properties [@chamberlain_soil_2018] and vegetation structure [@matthes_parsing_2014], and often exhibit substantial inter-annual variability due to disturbance and successional processes [@anderson_variation_2016; @wilson_multiyear_2016; @oikawa_evaluation_2017; @pugh_wetland_2018]. A better functional understanding wetland GHG exchange and responses to disturbance and management is required for a more robust assessment of the climate benefit of wetland restoration.

Understanding such responses can be challenging, as ecosystems are complex systems where non-linear, lagged, and feedback interactions are common [@ruddell_ecohydrologic_2009; @sturtevant_identifying_2016]. In some cases, findings from small-scale experiments may not be relevant to whole ecosystem dynamics [@schindler_dilemma_2012], which requires tools to delineate dominant or causal interactions from *in situ* observations. A number of statistical approaches common to neuroscience and econometrics, yet relatively novel to ecology, are suited to this task [@rinderer_assessing_2018], and are useful for constructing appropriate models to predict future ecosystem states [@larsen_appropriate_2016; @getz_wayne_m_making_2017]. These include non-parametric information theory metrics, such as mutual information (MI), that can identify dominant lagged, non-linear, and non-monotic relationships commonly observed in flux timeseries [@sturtevant_identifying_2016; @chamberlain_soil_2018; @knox_direct_2018]. Other metrics aimed at identifying causal relationships, such as Granger causality, have been used to identify causal relationships between GHG fluxes and the critical timescales at which they occur in forest [@detto_causality_2012] and rice ecosystems [@hatala_gross_2012]. Transfer entropy (TE), an additional information theory metric equivalent to Granger causality under Gaussian conditions [@barnett_granger_2009], has also been applied to successfully to understand functional changes in ecohydrologic systems during drought [@ruddell_ecohydrologic_2009; @ruddell_ecohydrologic_2009-1] and how C cycling responds to restoration practices [@larsen_disrupted_2017]. These statistical methods provide a means to better understand functional changes across disturbance events, which is critical to improving our understanding of how restored systems might respond to management or disturbance.

The objectives of this study were to determine the effect of drought-induced wetland salinization on CO~2~ and CH~4~ fluxes from restored wetlands in the Sacramento-San Joaquin Delta of California, USA. We measured ecosystem-scale CO~2~ and CH~4~ fluxes by eddy covariance from a restored wetland complex that experienced a substantial rise in salinity over the course of five years, and was then subsequently flushed with freshwater by land managers during the sixth year. We then used (1) information theory metrics to identify the how couplings and causal relationships between temperature, GEP, and CH~4~ fluxes change over the course of wetland salinization, and (2) used machine learning approaches to quantify the overall reduction in GEP and CH~4~ fluxes due to salinization. Understanding and predicting the response of Delta wetland GHG exchange to salinization events is important from a policy perspective, as wetland restoration in the region is funded through California's Cap-and-Trade Program [@deverel_implications_2017; @oikawa_evaluation_2017] and freshwater is projected to become an increasingly limited resource in the state's future [@vicuna_sebastian_sensitivity_2007].

# Materials and Methods

# Results
## Salinity effects
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat6_processing.R") #wetland experiencing salinity rise
source("R/peat19_processing.R") #baseline mature wetland w/o salinity rise
```

```{r fig1, fig.height = 8, fig.width = 7}
peat6_sub <- select(peat6_daily, -starts_with("DO"), -starts_with("CO", ignore.case = F))

both_sites <- rbind(peat19_daily, peat6_sub)

a <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=mgCH4, color=site)) +
  geom_point(size = 0.5)

b <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=gCO2, color=site)) +
  geom_point(size = 0.5)

c <- ggplot(subset(both_sites, year > 2013)) +
  geom_line(aes(x=datetime, y=gER, color=site)) #+
  #geom_line(aes(x=datetime, y=gGEP, color=site))

d <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=Tair, color=site)) +
  geom_line()

grid.arrange(a, b, c, d, ncol=1)
```

```{r fig2, fig.height=7, fig.width=3.5}
#Average monthly fluxes prior to salinity peak in 2016 
MBav <- subset(peat6_monthly, year > 2012 & year < 2016) %>%
  group_by(month) %>%
  summarise(CH4 = mean(gCH4),
            sdCH4 = 1.96*sd(gCH4)/sqrt(length(gCH4)),
            NEE = mean(gNEE),
            sdNEE = 1.96*sd(gNEE)/sqrt(length(gNEE)),
            ER = mean(gER),
            sdER = 1.96*sd(gER)/sqrt(length(gER)),
            GEP = mean(gGEP),
            sdGEP = 1.96*sd(gGEP)/sqrt(length(gGEP)),
            n = length(gCH4))

#Mean conductivity on Twitchell Island without salinity rise
Tw_av <- rbind(peat19_monthly, peat6_monthly) %>%
  group_by(month) %>%
  summarise(cond = mean(mCond, na.rm=T),
            sd.cond = 1.96*sd(mCond, na.rm=T)/sqrt(length(mCond)))


# plot theme
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.x = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

# four panels
a <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=CH4-sdCH4, ymax=CH4+sdCH4), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=CH4)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gCH4, color=factor(year))) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  ylab(expression(F[CH4]~" (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="a ", vjust=1, hjust=1, fontface = "bold") +
  time_series + 
  theme(legend.title=element_blank(), legend.margin = margin(0, 0, 0, 0),
        legend.text = element_text(size=8), legend.position=c(.15, .75),
        legend.background = element_rect(fill = "transparent"))

b <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=ER-sdER, ymax=ER+sdER), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=ER)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gER, color=factor(year))) +
  ylab(expression("ER (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="b ", vjust=1, hjust=1,fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

c <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=GEP-sdGEP, ymax=GEP+sdGEP), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=GEP)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=gGEP, color=factor(year))) +
  ylab(expression("GEP (g C"~m^{-2}~mo^{-1}~")")) + 
  annotate("text", x=Inf, y=Inf, label="c ", vjust=1, hjust=1, fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

d <- ggplot() +
  geom_ribbon(data=Tw_av, aes(x=month, ymin=cond-sd.cond, ymax=cond+sd.cond), fill="light grey") +
  geom_line(data=Tw_av, aes(x=month, y=cond)) +
  geom_line(data=subset(peat6_monthly, year > 2014), aes(x=month, y=mCond, color=factor(year))) +
  ylab("Conductivity (mS)")  + 
  annotate("text", x=Inf, y=Inf, label="d ", vjust=1, hjust=1, fontface = "bold") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  time_series + theme(legend.position = "none")

#Multiple plots aligned
g1 <- ggplotGrob(a)
g2 <- ggplotGrob(b)
g3 <- ggplotGrob(c)
g4 <- ggplotGrob(d)

grid.arrange(g1, g2, g3, g4, ncol = 1)
```

## Connectivity across time

```{r pca, echo = F}
subset <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD, Ts, Tw, PA, gER, Cond)

pca_test <- preProcess(subset, method = c("medianImpute"))
out_sub <- predict(pca_test, subset)

test <- prcomp(out_sub, scale. = T)

biplot(test)
```

```{r entropy_analysis_peat, fig.cap="Monthly relative mutual information (MI; top panel) shared between GEP and F~CH4~ (green line) as well as T~a~ and F~CH4~ (red line). All MI values are corrected for bias due to varied sample size using Monte Carlo reshuffling of time series such that MI = 0 represents the threshold of statistical significance (*P* < 0.05). Bottom panel displays daily mean water column conductivity (mS).", fig.width=7, fig.height=5, dpi=600}
source("R/MI_functions.R")
source("R/signficance_tests.R")

#create a growing month stamp
peat6_all$month <- month(peat6_all$datetime)
peat6_all$m_yr <- paste(peat6_all$year, peat6_all$month, sep = "")

#create data list by month
by_month <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(m_yr) %>%
  nest()

#plot that needs to be dealt with later
clipped_monthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_monthly$date <- as.Date(clipped_monthly$datetime)
clipped_monthly$GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm, data_list = by_month)
clipped_monthly$Tair_CH4_mi <- mi_timeseries(TA.x, wm, data_list = by_month)

# Monte Carlo significance thresholds (dont run everytime as they take a while)
sig_gpp <- conf_series(gpp_ANNnight, wm, data_list = by_month, runs = 100, alpha = 0.05)
sig_Tair <- conf_series(TA.x, wm, data_list = by_month, runs = 100, alpha = 0.05)

# clipped_monthly$Tair_ma <- na.ma(ma(clipped_monthly$Tair_ma, 3))
# clipped_monthly$GPP_ma <- na.ma(ma(clipped_monthly$GPP_ma, 3))

# Correct MI values of bias due to sample size, now anything over zero is significant at p < 0.01
clipped_monthly$TairMI_corrected <- clipped_monthly$Tair_CH4_mi - sig_Tair
clipped_monthly$gppMI_corrected <- clipped_monthly$GPP_CH4_mi - sig_gpp

a <- ggplot(clipped_monthly) +
  geom_point(aes(x = date, y = TairMI_corrected), color = "red", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(TairMI_corrected, 3)), color = "red", size = 1) +
  geom_point(aes(x = date, y = gppMI_corrected), color = "dark green", alpha = 0.5) +
  geom_line(aes(x = date, y = ma(gppMI_corrected, 3)), color = "dark green", size = 1) +  
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  ylab(expression(bolditalic(MI)~"(x,"~F[CH4]~")")) + theme_bw() + xlab("")

b <- ggplot(subset(peat6_daily, year > 2011 & year < 2018), aes(x=as.Date(datetime), y=Cond)) +
  geom_line(color = "blue") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y")) +
  ylab("Conductivity (mS)") + xlab("Year") +
  theme_bw()
  
grid.arrange(a, b, ncol=1)
```

```{r tr_GPP_to_CH4, fig.cap="Normalized transfer entropy (***T***) from (top panel) half-hourly GEP to F~CH4~ and (bottom panel) T~a~ to F~CH4~ over the growing season during years of low salinity (2012-2013) and high salinity (2015-2016). The red line represents significance threshold (*P* < 0.05) based on Monte Carlo reshuffling of the time series. Other lines represented ***T*** at various lags of GEP or T~a~ and color demonstrates mean conductivity (mS) during the respective year. Years 2014 and 2017 are omitted from analysis because other disturbance events affected wetlands processes and confound entropy analyses.", fig.height=4, fig.width=7.5, dpi=600}
source("R/TR_functions.R")
source("R/transformations.R")

#apply anomaly filter
anom_data <- peat6_all %>%
  arrange(year, time) %>%
  mutate(CH4_anom = anom_filter(wm_gf),
         TA_anom = anom_filter(TA.x),
         GPP_anom = anom_filter(gpp_ANNnight))

anom_data <- arrange(anom_data, datetime)
anom_data$CH4na_anom <- ifelse(is.na(anom_data$wm), NA, anom_data$CH4_anom) # re-insert missing values

# Focus only on high v. low salinity years without additional disturbance
# other disturbance includes the 2014 insect outbreak and 2017 flushing events
anom_data <- subset(anom_data, year %in% c(2012, 2013, 2015, 2016))

#create data list by annual growing season
by_year <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  nest()

#calculate lagged transfer entropy time series (up to 2 day lag at half-hourly timestep)
# for GPP -> CH4
tr_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 72, bins = 10, normalize = T)

mc_wm_gpp <- tr_timeseries(GPP_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 72, bins = 10, runs = 25, alpha = 0.05, normalize = T)
# for TA -> CH4
tr_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "observed", 
                           lags = 72, bins = 10, normalize = T)

mc_wm_ta <- tr_timeseries(TA_anom, CH4na_anom, data_list = by_year, type = "shuffled", 
                           lags = 72, bins = 10, runs = 25, alpha = 0.05, normalize = T)

#unlist data to tidy dataframe
gpp.wm_df <- tidy_list(tr_wm_gpp); gpp.wm_shuf <- tidy_list(mc_wm_gpp)
ta.wm_df <- tidy_list(tr_wm_ta); ta.wm_shuf <- tidy_list(mc_wm_ta)

#add mean annual conductivity to dataframes by year
mean_cond <- anom_data %>%
  filter(DOY > 100 & DOY < 300) %>%
  group_by(year) %>%
  summarise(`Cond. (mS)` = mean(Cond, na.rm = T))

gpp.wm_df <- merge(gpp.wm_df, mean_cond, all.x = T) 
ta.wm_df <- merge(ta.wm_df, mean_cond, all.x = T) 

a <- ggplot() +
  geom_line(data = gpp.wm_df, aes(x = hour, y = tr, color = `Cond. (mS)`)) +
  geom_line(data = gpp.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = seq(0, 48, by = 6)) +
  xlab("") + ylab(expression(bolditalic(T)~"("~GEP %->% F[CH4]~")")) +
  theme(legend.position = "none")

b <- ggplot() +
  geom_line(data = ta.wm_df, aes(x = hour, y = tr, color = `Cond. (mS)`)) +
  geom_line(data = ta.wm_shuf, aes(x = hour, y = tr), color = "red") +
  facet_wrap(~year, nrow = 1) +
  scale_x_continuous(breaks = seq(0, 48, by = 6)) +
  xlab("Lag (hours)") + ylab(expression(bolditalic(T)~"("~T[a] %->% F[CH4]~")")) +
  theme(legend.margin = margin(0, 0, 0, 0),
        legend.position=c(.31, .7), legend.key.size =  unit(0.12, "in"),
        legend.text = element_text(size=8), legend.title = element_text(size=8.5, face = "bold"),
        legend.background = element_rect(fill = "transparent"))


#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); 

grid.arrange(iA, iB, ncol=1)
```

```{r pre_salinity_models, fig.cap="", fig.height=6, fig.width=3.5, dpi=600}
# train nn model on pre-salinity rise years and use to calculate reduction in component fluxes 
pre_rise <- peat6_daily %>%
  filter(year > 2011 & year < 2015) %>%
  select(mgCH4, gGEP, gER, mET, PAR, H, Tair, Tw, Ts, mVPD, PA, u., Rain, WTD)

# Random forest model for GEP trained pre-salinity rise
pre_GEP <- train(gGEP ~ Tair + PAR + mVPD + PA + u. + WTD, data = pre_rise,
                      method = "rf",
                      trControl = trainControl(method = "cv", number = 5),
                      preProcess = c("medianImpute"),
                      #tuneGrid = expand.grid(.decay = c(.1, 0.05, .01), .size = c(3, 5, 7, 9)),
                      na.action = na.pass,
                      linout = T)

# Random forest model for FCH4 trained pre-salinity rise
pre_CH4 <- train(mgCH4 ~ gGEP + Tair + mVPD + PA + u. + WTD, data = pre_rise,
                      method = "rf",
                      trControl = trainControl(method = "cv", number = 5),
                      preProcess = c("medianImpute"),
                      #tuneGrid = expand.grid(.decay = c(.1, 0.05, .01), .size = c(3, 5, 7, 9)),
                      na.action = na.pass)

# generate predictions for high salinity years for GEP and FCH4
pre_rise$GEPpred <- predict.train(pre_GEP, pre_rise, na.action = na.pass)
pre_rise$CH4pred <- predict.train(pre_CH4, pre_rise, na.action = na.pass)

# save ob v. pred summary statistics
GEPregr <- summary(lm(gGEP ~ GEPpred, data = pre_rise))
CH4regr <- summary(lm(mgCH4 ~ CH4pred, data = pre_rise))

a <- ggplot(pre_rise, aes(x = CH4pred, y = mgCH4)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1) +
  xlab(expression(F[CH4-Pred.]~" (mg C"~m^{-2}~d^{-1}~")")) +
  ylab(expression(F[CH4-Obs.]~" (mg C"~m^{-2}~d^{-1}~")")) +
  ggtitle("Pre-salinity rise (2012-2014)")

b <- ggplot(pre_rise, aes(x = GEPpred*-1, y = gGEP*-1)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1) +
  ylab(expression(GEP[Obs.]~"("~g~C~m^{-2}~d^{-1}~")")) +
  xlab(expression(GEP[Pred.]~"("~g~C~m^{-2}~d^{-1}~")"))

grid.arrange(a, b, ncol = 1)
```

```{r salinity_rise_predictions, fig.width=7.5, fig.height=4}
# extract data to make predictions on the salinity rise years
rise_years <- peat6_daily %>%
  filter(year > 2014 & year < 2017) %>%
  select(datetime, year, mgCH4, gGEP, gER, mET, PAR, H, Tair, Tw, Ts, mVPD, PA, u., Rain, WTD)

# generate predictions for high salinity years for GEP and FCH4
rise_years$GEPpred <- predict.train(pre_GEP, rise_years, na.action = na.pass)
rise_years$CH4pred <- predict.train(pre_CH4, rise_years, na.action = na.pass)

#plotting all the measures
a <- ggplot(rise_years) +
  geom_line(aes(x=datetime, y=mgCH4)) +
  geom_line(aes(x=datetime, y=CH4pred), color = "red") +
  ylab(expression(F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) + xlab("") +
  ggtitle("High salinity period (2015-2016)") +
  theme_classic()

b <- ggplot(rise_years) +
  geom_line(aes(x=datetime, y=gGEP*-1)) +
  geom_line(aes(x=datetime, y=GEPpred*-1), color = "red") +
  ylab(expression("GEP "~"("~g~C~m^{-2}~d^{-1}~")")) + xlab("Date") +
  theme_classic()

grid.arrange(a, b, ncol = 1)
```

```{r flux_reduction_estimates}
#calculate annual budgets for RF predictions and observations during salinity years
annual_budgets <- rise_years %>%
  group_by(year) %>%
  summarise(CH4yr = mean(mgCH4)*365/1000,
            CH4yr_rf = mean(CH4pred)*365/1000,
            GEPyr = mean(gGEP)*365,
            GEPyr_rf = mean(GEPpred)*365) %>%
  mutate(CH4pc = (CH4yr - CH4yr_rf)*100/CH4yr,
         GEPpc = (GEPyr - GEPyr_rf)*100/GEPyr)
```

<!-- ## Water chemistry -->

<!-- ```{r fig4, echo = F} -->
<!-- only16 <- subset(peat6_daily, year > 2015) -->
<!-- only16$jday <- lubridate::yday(only16$datetime) -->

<!-- only16$season <- ifelse(only16$jday %in% c(0:100) | only16$jday %in% c(300:365), "fallow",  -->
<!--                            "growing") -->

<!-- # ggplot(only16, aes(x=datetime)) + -->
<!-- #   geom_line(aes(y = DOrange1), color="red") + -->
<!-- #   geom_line(aes(y = DOdiff), color="blue") -->
<!-- #  -->
<!-- # ggplot(only16, aes(x=mgCH4, y=DOrange1, color=season)) + -->
<!-- #   geom_point() -->

<!-- regr <- lm(DOrange1 ~ mgCH4, data = subset(only16, season == "growing")); summary(regr) -->
<!-- ``` -->

```{r test, echo = F}
#CH4sub <- peat6_all[!is.na(peat6_all$wm), ]

by_year <- peat6_daily %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(year) %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  nest()

net_model <- function(df) {
  train(mgCH4 ~ ., data = df, method = "glmnet",
                 trControl = trainControl(method = "cv"),
                 preProcess = c("center", "scale", "medianImpute", "BoxCox"),
                 tuneGrid = expand.grid(.alpha = 1, .lambda = 0.1),
                 na.action = na.pass)
}

# hh_model <- function(df) {
#   train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD,
#         data = df,
#         method = "glmnet",
#         preProcess = c("center", "scale", "medianImpute", "BoxCox"),
#         na.action = na.pass)
# }

# subset <- peat6_daily %>%
#   ungroup() %>%
#   select(mgCH4, gGEP, gER, mET, H, Tair, Tw, mVPD, u., Rain, WTD) %>%
#   na.omit()

# #caret approach
# glm_reg <- train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD, 
#                  data = CH4sub, method = "glmnet",
#                  trControl = trainControl(method = "cv"),
#                  preProcess = c("center", "scale", "medianImpute"),
#                  #tuneGrid = expand.grid(.alpha = 1, .lambda = c(0, 0.001, 0.1, 1, 5, 10, 20, 50)),
#                  na.action = na.pass)

by_year <- by_year %>%
  mutate(model = map(data, net_model),
         Imp = map(model, varImp))

impo_list <- by_year$Imp

a <- impo_list[[1]]$importance
a$year <- 2012
a$vars <- rownames(a)

b <- impo_list[[2]]$importance
b$year <- 2013
b$vars <- rownames(b)

c <- impo_list[[3]]$importance
c$year <- 2014
c$vars <- rownames(c)

d <- impo_list[[4]]$importance
d$year <- 2015
d$vars <- rownames(d)

e <- impo_list[[5]]$importance
e$year <- 2016
e$vars <- rownames(e)

f <- impo_list[[6]]$importance
f$year <- 2017
f$vars <- rownames(f)

all_now <- do.call("rbind", list(a, b, c, d, e, f))

ggplot(all_now, aes(x=year, y=Overall, color=vars)) +
  #stat_smooth(se=F) +
  geom_point(size = 2) +
  geom_line()
```

# Discussion

# Acknowledgements
This research was supported in part by the U.S. Department of Energy’s Office of Science and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

