---
title: "Salinity study"
author: Samuel D Chamberlain^1^, Elke Eichelmann^1^, Kyle S Hemes^1^, Daphne J Szutu^1^,
  Joseph G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
link-citations: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
fontsize: 12pt
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
library(caret)
library(modelr)
library(forecast)
library(entropy)
```

<!-- \newpage -->

<!-- # Abstract -->

<!-- # Introduction -->

<!-- # Materials and Methods -->

<!-- # Results -->
## Salinity effects
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat6_processing.R") #wetland experiencing salinity rise
source("R/peat19_processing.R") #baseline mature wetland w/o salinity rise
```

```{r fig1, fig.height = 8, fig.width = 7}
peat6_sub <- select(peat6_daily, -starts_with("DO"), -starts_with("CO", ignore.case = F))

both_sites <- rbind(peat19_daily, peat6_sub)

a <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=mgCH4, color=site)) +
  geom_point(size = 0.5)

b <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=gCO2, color=site)) +
  geom_point(size = 0.5)

c <- ggplot(subset(both_sites, year > 2013)) +
  geom_line(aes(x=datetime, y=gER, color=site)) #+
  #geom_line(aes(x=datetime, y=gGEP, color=site))

d <- ggplot(subset(both_sites, year > 2013), aes(x=datetime, y=Tair, color=site)) +
  geom_line()

grid.arrange(a, b, c, d, ncol=1)
```

```{r fig2, fig.height=7, fig.width=4}
#Average monthly fluxes prior to salinity peak in 2016 
MBav <- subset(peat6_monthly, year > 2012 & year < 2016) %>%
  group_by(month) %>%
  summarise(CH4 = mean(gCH4),
            sdCH4 = 1.96*sd(gCH4)/sqrt(length(gCH4)),
            NEE = mean(gNEE),
            sdNEE = 1.96*sd(gNEE)/sqrt(length(gNEE)),
            ER = mean(gER),
            sdER = 1.96*sd(gER)/sqrt(length(gER)),
            GEP = mean(gGEP),
            sdGEP = 1.96*sd(gGEP)/sqrt(length(gGEP)),
            n = length(gCH4))

#Mean conductivity on Twitchell Island without salinity rise
Tw_av <- rbind(peat19_monthly, peat6_monthly) %>%
  group_by(month) %>%
  summarise(sal = mean(mCond, na.rm=T),
            sdsal = 1.96*sd(mCond, na.rm=T)/sqrt(length(mCond)))


a <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=CH4-sdCH4, ymax=CH4+sdCH4), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=CH4)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gCH4), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gCH4), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gCH4), color="#2c7bb6") +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  ylab(expression(F[CH4]~" (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="a ", vjust=1, hjust=1) +
  theme_bw() + theme(axis.title.x=element_blank())

b <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=ER-sdER, ymax=ER+sdER), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=ER)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gER), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gER), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gER), color="#2c7bb6") +
  ylab(expression("ER (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="b ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw() + theme(axis.title.x=element_blank())

c <- ggplot() +
  geom_ribbon(data=MBav, aes(x=month, ymin=GEP-sdGEP, ymax=GEP+sdGEP), fill="light grey") +
  geom_line(data=MBav, aes(x=month, y=GEP)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=gGEP), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=gGEP), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=gGEP), color="#2c7bb6") +
  ylab(expression("GEP (g C"~m^{-2}~mo^{-1}~")")) + annotate("text", x=Inf, y=Inf, label="c ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw() + theme(axis.title.x=element_blank())

d <- ggplot() +
  geom_ribbon(data=Tw_av, aes(x=month, ymin=sal-sdsal, ymax=sal+sdsal), fill="light grey") +
  geom_line(data=Tw_av, aes(x=month, y=sal)) +
  geom_line(data=subset(peat6_monthly, year==2015), aes(x=month, y=mCond), color="#fdae61") +
  geom_line(data=subset(peat6_monthly, year==2016), aes(x=month, y=mCond), color="#d7191c") +
  geom_line(data=subset(peat6_monthly, year==2017), aes(x=month, y=mCond), color="#2c7bb6") +
  ylab("Salinity")  + annotate("text", x=Inf, y=Inf, label="d ", vjust=1, hjust=1) +
  scale_x_continuous(name="Month", breaks=seq(0,12,2)) +
  theme_bw()

#Multiple plots aligned
g1 <- ggplotGrob(a)
g2 <- ggplotGrob(b)
g3 <- ggplotGrob(c)
g4 <- ggplotGrob(d)
g <- rbind(g1, g2, g3, g4, size="first");

grid.arrange(g)
```

## Models by year for comparison
```{r anomaly_filter, echo = F, message = F}
anom_filter <- function(var, window = 48*7) {
  var - ma(var, order = window)
}

test <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  map_df(anom_filter)
  
#cor(test, method = "spearman", use = "complete.obs")

```

```{r pca, echo = F}
subset <- peat6_daily %>%
  ungroup() %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD, Ts, Tw, PA, gER, Cond)

pca_test <- preProcess(subset, method = c("medianImpute"))
out_sub <- predict(pca_test, subset)

test <- prcomp(out_sub, scale. = T)

biplot(test)
```

```{r entropy_analysis_peat, fig.width = 7, fig.height= 8}
source("R/MI_functions.R")

# # test multiple binning strategies
# bin_v = seq(4, 22, by = 2)
# par(mfrow=c(2, 5))
# 
# for (i in bin_v) {
#   a <- lag_mi(peat6_all$ustar, peat6_all$wm, lags = 48, bins = i)
#   plot(a)
# }

#create a growing month stamp
peat6_all$month <- month(peat6_all$datetime)
peat6_all$m_yr <- paste(peat6_all$year, peat6_all$month, sep = "")

#shuffle data for comparison w/ monthly gap magnitude retained
peat6_all <- peat6_all %>%
  group_by(month) %>%
  mutate(H_shuf = sample(TA.x, length(TA.x), replace = F),
         wm_shuf = sample(wm, length(wm), replace = F))

#create data list by month
by_month <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(m_yr) %>%
  nest()

#plot that needs to be dealt with later
clipped_monthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_monthly$GPP_CH4_mi <- mi_timeseries(gpp_ANNnight, wm, data_list = by_month)
clipped_monthly$ET_CH4_mi <- mi_timeseries(H_shuf, wm_shuf, data_list = by_month)
clipped_monthly$Tair_CH4_mi <- mi_timeseries(TA.x, wm, data_list = by_month)

# # Monte Carlo significance thresholds (dont run everytime as they take a while)
# #sig_gpp <- conf_series(gpp_ANNnight, wm, data_list = by_month, runs = 1000)
# # mean sig_gpp across all months: 0.02311099
# 
# #sig_Tair <- conf_series(TA.x, wm, data_list = by_month, runs = 1000)
# # mean sig_Tair across all months: 0.03119986
#
# #sig_H <- conf_series(H_gf, wm, data_list = by_month, runs = 100)
# # mean sig_Tair across all months: 0.03119986

a <- ggplot(clipped_monthly) +
  geom_hline(yintercept = 0.02311099*2.36, size = 0.5, color = "dark green", linetype = "dashed") +
  geom_hline(yintercept = 0.03119986*2.36, size = 0.5, color = "red", linetype = "dashed") +
  #geom_hline(yintercept = mean(sig_H)*2.36, size = 0.5, color = "blue", linetype = "dashed") +
  geom_point(aes(x = datetime, y = Tair_CH4_mi), color = "red", alpha = 0.5) +
  geom_line(aes(x = datetime, y = ma(Tair_CH4_mi, 3)), color = "red", size = 1) +
  geom_point(aes(x = datetime, y = GPP_CH4_mi), color = "dark green", alpha = 0.5) +
  geom_line(aes(x = datetime, y = ma(GPP_CH4_mi, 3)), color = "dark green", size = 1) +
  geom_point(aes(x = datetime, y = ET_CH4_mi), color = "blue", alpha = 0.5) +
  geom_line(aes(x = datetime, y = ma(ET_CH4_mi, 3)), color = "blue", size = 1) +
  ylab("Mutual information") + theme_bw() + xlab("")

b <- ggplot(subset(peat6_daily, year > 2011 & year < 2018), aes(x=datetime, y=Cond)) +
  geom_line(color = "blue") +
  ylab("Conductivity (mS)") + theme_bw()
  
grid.arrange(a, b, ncol=1)
```

```{r bias_by_samples, fig.height= 5, fig.width=6}
#use monthly from above and compare to two month chunks

# create a bimonthly grouping
peat6_all$bimonth <- paste(peat6_all$year, floor((peat6_all$month+1)/2), sep="")

by_bimonth <- peat6_all %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(bimonth) %>%
  nest()

#reshuffle by gap magnitude preserved
peat6_all <- peat6_all %>%
  group_by(bimonth) %>%
  mutate(H_shuf = sample(TA.x, length(TA.x), replace = F),
         wm_shuf = sample(wm, length(wm), replace = F))

#calculate bimonthly MI
clipped_bimonthly <- subset(peat6_monthly, year > 2011 & year < 2018)
clipped_bimonthly$bimonth <- paste(clipped_bimonthly$year, floor((clipped_bimonthly$month+1)/2), sep="")
clipped_bimonthly <- aggregate(clipped_bimonthly, by=list(clipped_bimonthly$bimonth), FUN=mean)

clipped_bimonthly$ET_CH4_mi <- mi_timeseries(H_shuf, wm_shuf, data_list = by_bimonth)

ggplot() +
  geom_point(data = clipped_monthly, aes(x = CH4_obs, y = ET_CH4_mi)) +
  geom_point(data = clipped_bimonthly, aes(x = CH4_obs, y = ET_CH4_mi), color = "red") +
  xlab("Number of flux obs.") + ylab("MI (reshuffled data)")
```

```{r test, echo = F}
#CH4sub <- peat6_all[!is.na(peat6_all$wm), ]

by_year <- peat6_daily %>%
  filter(year > 2011 & year < 2018) %>%
  group_by(year) %>%
  select(mgCH4, gGEP, mET, H, Tair, mVPD, u., Rain, WTD) %>%
  nest()

net_model <- function(df) {
  train(mgCH4 ~ ., data = df, method = "glmnet",
                 trControl = trainControl(method = "cv"),
                 preProcess = c("center", "scale", "medianImpute", "BoxCox"),
                 tuneGrid = expand.grid(.alpha = 1, .lambda = 0.1),
                 na.action = na.pass)
}

# hh_model <- function(df) {
#   train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD,
#         data = df,
#         method = "glmnet",
#         preProcess = c("center", "scale", "medianImpute", "BoxCox"),
#         na.action = na.pass)
# }

# subset <- peat6_daily %>%
#   ungroup() %>%
#   select(mgCH4, gGEP, gER, mET, H, Tair, Tw, mVPD, u., Rain, WTD) %>%
#   na.omit()

# #caret approach
# glm_reg <- train(wm ~ gpp_ANNnight + wq_gf + wt_gf + TA.y + VPD + ustar + PRECIP + WT + WD, 
#                  data = CH4sub, method = "glmnet",
#                  trControl = trainControl(method = "cv"),
#                  preProcess = c("center", "scale", "medianImpute"),
#                  #tuneGrid = expand.grid(.alpha = 1, .lambda = c(0, 0.001, 0.1, 1, 5, 10, 20, 50)),
#                  na.action = na.pass)

by_year <- by_year %>%
  mutate(model = map(data, net_model),
         Imp = map(model, varImp))

impo_list <- by_year$Imp

a <- impo_list[[1]]$importance
a$year <- 2012
a$vars <- rownames(a)

b <- impo_list[[2]]$importance
b$year <- 2013
b$vars <- rownames(b)

c <- impo_list[[3]]$importance
c$year <- 2014
c$vars <- rownames(c)

d <- impo_list[[4]]$importance
d$year <- 2015
d$vars <- rownames(d)

e <- impo_list[[5]]$importance
e$year <- 2016
e$vars <- rownames(e)

f <- impo_list[[6]]$importance
f$year <- 2017
f$vars <- rownames(f)

all_now <- do.call("rbind", list(a, b, c, d, e, f))

# ggplot(all_now, aes(x=year, y=Overall, color=vars)) +
#   #stat_smooth(se=F) +
#   geom_point(size = 2) +
#   geom_line()

# ggplot(all_now, aes(x = year, y = vars)) +
#   geom_tile(aes(fill = Overall))
```

<!-- ## Information theory approach -->


<!-- ## Water chemistry -->

<!-- ```{r fig4, echo = F} -->
<!-- only16 <- subset(peat6_daily, year > 2015) -->
<!-- only16$jday <- lubridate::yday(only16$datetime) -->

<!-- only16$season <- ifelse(only16$jday %in% c(0:100) | only16$jday %in% c(300:365), "fallow",  -->
<!--                            "growing") -->

<!-- # ggplot(only16, aes(x=datetime)) + -->
<!-- #   geom_line(aes(y = DOrange1), color="red") + -->
<!-- #   geom_line(aes(y = DOdiff), color="blue") -->
<!-- #  -->
<!-- # ggplot(only16, aes(x=mgCH4, y=DOrange1, color=season)) + -->
<!-- #   geom_point() -->

<!-- regr <- lm(DOrange1 ~ mgCH4, data = subset(only16, season == "growing")); summary(regr) -->
<!-- ``` -->

<!-- # Discussion -->

# Acknowledgements
This research was supported in part by the U.S. Department of Energy’s Office of Science and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

